---
title: "Littoral habitat structure influences freshwater crayfish populations"
author: "Olivier V. Raven"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
execute:
  freeze: auto
  warning: false
  message: false
---

## Overview
This site accompanies the scientific work on: *Littoral habitat structure influences freshwater crayfish populations in five volcanic lakes of Aotearoa New Zealand*.

- **Code repository:** https://github.com/OlivierRaven/Koura_shoreline_habitats.git
- **Data DOI:** (add DOI link)
- **Paper DOI:** (add DOI link)
- **Quartro manuscript** https://olivierraven.github.io/Koura_shoreline_habitats/
- **Title:** Littoral habitat structure influences freshwater crayfish populations in five volcanic lakes of Aotearoa New Zealand

## Setup
```{r}
#| label: setup and load-data
knitr::opts_knit$set(root.dir = normalizePath(".."))
options(warn = -1)
packages <- c(
  "here","kableExtra","emmeans","rstatix","multcompView","brglm2","patchwork","gratia","pROC","caret",
  "mgcv","tidyr","purrr","corrplot","glmmTMB","performance","tibble","tidyverse",
  "dplyr","writexl","readxl","lubridate")

quiet_load <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    suppressWarnings(suppressMessages(install.packages(pkg, dependencies = TRUE)))
  }
  suppressPackageStartupMessages(require(pkg, character.only = TRUE, quietly = TRUE))
  invisible(TRUE)
}

invisible(lapply(packages, quiet_load))


exc_file_dir <- "data/raw/Natural_habitat.xlsx"
der_data_dir <- "data/derived"
out_dir      <- "outputs"

Site_info          <- read_excel(exc_file_dir, sheet = "Site_info")
Monitoring_data    <- read_excel(exc_file_dir, sheet = "Monitoring_data")
Weed_data          <- read_excel(exc_file_dir, sheet = "Weed_data")  %>% dplyr::select(-starts_with("..."))
Fish_data          <- read_excel(exc_file_dir, sheet = "Fish_data")  %>% dplyr::select(-starts_with("..."))
Macroinvertebrates <- read_excel(exc_file_dir, sheet = "Macroinvertebrates")

Site_info          <- Site_info          |> dplyr::filter(!stringr::str_ends(Monitoring_ID, "_1"))
Monitoring_data    <- Monitoring_data    |> dplyr::filter(!stringr::str_ends(Monitoring_ID, "_1"))
Weed_data          <- Weed_data          |> dplyr::filter(!stringr::str_ends(Monitoring_ID, "_1"))
Fish_data          <- Fish_data          |> dplyr::filter(!stringr::str_ends(Monitoring_ID, "_1"))
Macroinvertebrates <- Macroinvertebrates |> dplyr::filter(!stringr::str_ends(Monitoring_ID, "_1"))

# Save as csv to derived
write.csv(Site_info,          file.path(der_data_dir, "Site_info.csv"),          row.names = FALSE)
write.csv(Monitoring_data,    file.path(der_data_dir, "Monitoring_data.csv"),    row.names = FALSE)
write.csv(Weed_data,          file.path(der_data_dir, "Weed_data.csv"),          row.names = FALSE)
write.csv(Fish_data,          file.path(der_data_dir, "Fish_data.csv"),          row.names = FALSE)
write.csv(Macroinvertebrates, file.path(der_data_dir, "Macroinvertebrates.csv"), row.names = FALSE)

# Reproducibility note 
# The sections above require data/raw/Natural_habitat.xlsx (not tracked).
# To reproduce analyses and figures without the raw Excel file,
# run from here using the derived CSVs in data/derived/:

Site_info          <- read.csv(file.path(der_data_dir, "Site_info.csv"))
Monitoring_data    <- read.csv(file.path(der_data_dir, "Monitoring_data.csv"))
Weed_data          <- read.csv(file.path(der_data_dir, "Weed_data.csv"))
Fish_data          <- read.csv(file.path(der_data_dir, "Fish_data.csv"))
Macroinvertebrates <- read.csv(file.path(der_data_dir, "Macroinvertebrates.csv"))

# Set plot theme
base_theme_bw <- theme_classic() +
  theme(
    text = element_text(family = "Arial", size = 8),
    axis.title = element_text(face = "plain"),
    axis.text = element_text(face = "plain"),
    plot.title = element_text(face = "plain"),
    strip.text = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3)
  )

theme_set(base_theme_bw)

clean_smooth_title <- function(x) {
  x <- gsub("^s\\((.*)\\)$", "\\1", x)
  x <- gsub("_", " ", x)
  x
}

```


# CPUE and BPUE derivation
## Length–weight model and predicted weights
```{r}
#| label: length-weight
#| fig-width: 6
#| fig-height: 4
#| dpi: 120
#| fig-cap: "Fig. S1 Length–weight relationship used to estimate missing kōura weights. Observed (measured) and model-predicted body weights are shown in relation to orbital carapace length (OCL). The log₁₀–log₁₀ regression was fitted to individuals with both length and weight measurements (n = 240) and used to estimate missing weights (n = 81). Predicted values were back-transformed using a lognormal bias correction factor. The fitted regression equation is shown on the plot."

lw_model <- lm(
log10(Weight_g) ~ log10(Length_mm),
data = Fish_data,
subset = Species_name == "Freshwater_crayfish" | Maori_name == "Kōura",
na.action = na.exclude
)

sigma_log10 <- sigma(lw_model)
c <- 10^(0.5 * sigma_log10^2)

Fish_data <- Fish_data |>
mutate(
is_koura = Species_name == "Freshwater_crayfish" | Maori_name == "Kōura",
Predicted_weight = case_when(
!is.na(Weight_g) ~ Weight_g,
is_koura ~ {
pred_log <- predict(lw_model, newdata = pick(everything()))
10^(pred_log) * c
},
TRUE ~ NA_real_
),
Weight_source = case_when(
is_koura & !is.na(Weight_g) ~ "measured",
is_koura & is.na(Weight_g) ~ "predicted",
TRUE ~ NA_character_
)
)

a <- coef(lw_model)[1]
b <- coef(lw_model)[2]
formula_text <- paste0(
"log10(Weight[g]) = ", round(a, 3), " + ", round(b, 3),
" * log10(OCL Length[mm])"
)

length_weight_plot <- ggplot(
Fish_data |> dplyr::filter(is_koura),
aes(x = Length_mm, y = Predicted_weight, shape = Weight_source)) +
geom_point(size = 2.2, colour = "black", stroke = 0.6) +
scale_shape_manual(values = c("measured" = 16, "predicted" = 1)) +
labs(x = "OCL length (mm)", y = "Weight (g)", shape = "Weight source") +
annotate("text", x = Inf, y = Inf, label = formula_text, hjust = 1.1, vjust = 1.3, size = 3) 

ggsave(filename = file.path(out_dir, "length_weight_plot.png"),plot = length_weight_plot,width = 170, height = 120, units = "mm", dpi = 1200)

length_weight_plot
```

## CPUE and BPUE calculations
```{r}
#| label: cpue-bpue

CPUE_BPUE_legacy <- Fish_data %>%
filter(!is.na(Species)) %>%
group_by(Monitoring_ID, Species, Net_type) %>%
dplyr::reframe(
Total_Individuals = sum(Amount, na.rm = TRUE),
Total_Weight      = sum(Predicted_weight, na.rm = TRUE),
Total_Effort      = dplyr::first(Amount_nets),
CPUE              = Total_Individuals / Total_Effort,
BPUE              = Total_Weight / Total_Effort,
Mean_Length       = mean(Length_mm, na.rm = TRUE),
Min_Length        = ifelse(all(is.na(Length_mm)), NA, min(Length_mm, na.rm = TRUE)),
Max_Length        = ifelse(all(is.na(Length_mm)), NA, max(Length_mm, na.rm = TRUE)),
Mean_Weight       = mean(Predicted_weight, na.rm = TRUE),
Min_Weight        = ifelse(all(is.na(Predicted_weight)), NA, min(Predicted_weight, na.rm = TRUE)),
Max_Weight        = ifelse(all(is.na(Predicted_weight)), NA, max(Predicted_weight, na.rm = TRUE))
)

CPUE_BPUE_weighted <- CPUE_BPUE_legacy %>%
group_by(Monitoring_ID, Species) %>%
summarise(
Total_Individuals       = sum(Total_Individuals, na.rm = TRUE),
Total_Weight            = sum(Total_Weight,      na.rm = TRUE),
Weighted_CPUE_numerator = sum(CPUE * Total_Effort,  na.rm = TRUE),
Weighted_BPUE_numerator = sum(BPUE * Total_Effort,  na.rm = TRUE),
Total_Effort_sum        = sum(Total_Effort,         na.rm = TRUE),
Mean_Length             = mean(Mean_Length, na.rm = TRUE),
Min_Length              = ifelse(all(is.na(Min_Length)), NA, min(Min_Length, na.rm = TRUE)),
Max_Length              = ifelse(all(is.na(Max_Length)), NA, max(Max_Length, na.rm = TRUE)),
Mean_Weight             = mean(Mean_Weight, na.rm = TRUE),
Min_Weight              = ifelse(all(is.na(Min_Weight)), NA, min(Min_Weight, na.rm = TRUE)),
Max_Weight              = ifelse(all(is.na(Max_Weight)), NA, max(Max_Weight, na.rm = TRUE)),
.groups = "drop"
) %>%
mutate(
Total_Effort_sum = ifelse(Monitoring_ID %in% c("96_0", "101_0", "117_1", "119_1"), 3, 4),
Weighted_CPUE    = Weighted_CPUE_numerator / Total_Effort_sum,
Weighted_BPUE    = Weighted_BPUE_numerator / Total_Effort_sum
)

species_presence_absence <- Fish_data %>%
filter(!is.na(Species)) %>%
distinct(Monitoring_ID, Species) %>%
mutate(Presence = 1) %>%
pivot_wider(
names_from  = Species,
values_from = Presence,
values_fill = list(Presence = 0),
names_prefix = "Presence_"
) %>%
mutate(Predator_Fish_Presence = pmax(Presence_Trout, Presence_Eel, Presence_Catfish))

CPUE_BPUE_weighted_summary <- CPUE_BPUE_weighted %>%
pivot_wider(
names_from  = Species,
values_from = c(
Total_Individuals, Weighted_CPUE, Weighted_BPUE, Total_Weight,
Mean_Length, Mean_Weight, Weighted_CPUE_numerator, Weighted_BPUE_numerator,
Total_Effort_sum, Min_Length, Max_Length, Min_Weight, Max_Weight
),
names_sep   = "_",
values_fill = list(Total_Individuals = 0, Weighted_CPUE = 0, Weighted_BPUE = 0)
) %>%
mutate(
Richness  = rowSums(dplyr::select(., starts_with("Total_Individuals_")) > 0),
Abundance = rowSums(dplyr::select(., starts_with("Total_Individuals_") & !ends_with(c("_Bullies", "_Common_smelt"))))
)

```

# Combined dataset and habitat classification
```{r}
#| label: combine-data

unit_metadata <- Monitoring_data %>%
dplyr::select(Parameter, Unit) %>%
distinct()

Monitoring_summary <- Monitoring_data %>%
dplyr::select(-Group, -Notes, -Unit) %>%
pivot_wider(
names_from  = c(Parameter),
values_from = Value,
values_fill = list(Value = NA)
) %>%
mutate(across(
c(
Bottom_visible, Water_clarity, Depth_10m, Slope, Riparian_vegetation, Vegetation_nearby,
Overhanging_trees, Erosion, Structure, Bedrock, Boulders, Cobble, Gravel,
Sand, Mud, Organic_matter, Rock_size, Temperature, DO_mgl, DO_percent,
Conductivity, Specific_conductivity, pH, Wood_cover
),
~ as.numeric(.)
))

Weed_summary <- Weed_data %>%
group_by(Monitoring_ID, Weed_Type, Native_Status) %>%
summarise(Total_Cover = sum(Percentage_Cover, na.rm = TRUE), .groups = "drop") %>%
pivot_wider(
names_from  = c(Weed_Type, Native_Status),
values_from = Total_Cover,
values_fill = 0
)

Macroinvertebrates_sum <- Macroinvertebrates %>%
group_by(Monitoring_ID, Species) %>%
summarise(Total_amount = sum(Amount, na.rm = TRUE), .groups = "drop") %>%
pivot_wider(names_from = c(Species), values_from = Total_amount, values_fill = 0) %>%
mutate(
Invertebrates_Richness  = rowSums(dplyr::select(., -Monitoring_ID) > 0),
Invertebrates_Abundance = rowSums(dplyr::select(., -Monitoring_ID))
)

Monitoring_CPUE_data <- Site_info %>%
left_join(Monitoring_summary, by = "Monitoring_ID") %>%
left_join(
Weed_summary %>% dplyr::select(
Monitoring_ID,
Emergent_Native, Emergent_Non_Native,
Submerged_Native, Submerged_Non_Native, Turf_Native
),
by = "Monitoring_ID"
) %>%
left_join(CPUE_BPUE_weighted_summary, by = "Monitoring_ID") %>%
left_join(species_presence_absence, by = "Monitoring_ID") %>%
left_join(Macroinvertebrates_sum, by = "Monitoring_ID")

Monitoring_CPUE_data <- Monitoring_CPUE_data %>%
  mutate(
    Presence_rocks = if_else(Cobble > 1 | Boulders > 1, 1, 0),
    Slope_5m       = 5 / Distance_5m,
    Site_ID_       = Site_ID.x - 60,
    Monitoring_ID_ = paste0(
      as.numeric(sub("_.*", "", Monitoring_ID)) - 60,
      sub("^[^_]*", "", Monitoring_ID)
    ),
    Monitoring     = sub(".*?_", "", Monitoring_ID),
    Date           = as.Date(Date_Time),
    Time           = format(as.POSIXct(Date_Time), "%H:%M:%S"),
    Year           = lubridate::year(Date_Time),
    Month          = lubridate::month(Date_Time, label = TRUE),
    Day            = lubridate::day(Date_Time),
    Season         = case_when(
      Month %in% c("Dec", "Jan", "Feb") ~ "Summer",
      Month %in% c("Mar", "Apr", "May") ~ "Autumn",
      Month %in% c("Jun", "Jul", "Aug") ~ "Winter",
      Month %in% c("Sep", "Oct", "Nov") ~ "Spring",
      TRUE ~ NA_character_
    ),
    Date_Time_Numeric = as.numeric(Date_Time)
  )

habitat_classification <- Monitoring_CPUE_data %>%
dplyr::select(
Monitoring_ID, DHT, Lake,
Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter,
Emergent_Native, Emergent_Non_Native,
Submerged_Native, Submerged_Non_Native, Wood_cover
) %>%
pivot_longer(
cols = c(
Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter,
Emergent_Native, Emergent_Non_Native,
Submerged_Native, Submerged_Non_Native, Wood_cover
),
names_to = "Type",
values_to = "Percentage"
) %>%
group_by(Monitoring_ID) %>%
summarise(
Rocky_Percentage = sum(Percentage[Type %in% c("Bedrock", "Boulders", "Cobble")], na.rm = TRUE),
Sand_Percentage  = sum(Percentage[Type == "Sand"], na.rm = TRUE),
Mud_Percentage   = sum(Percentage[Type %in% c("Mud", "Organic_matter")], na.rm = TRUE),
Emergent_Percentage = sum(Percentage[Type == "Emergent_Native"], na.rm = TRUE),
Substrate_index = sum(
0.08 * Percentage[Type == "Bedrock"] +
0.07 * Percentage[Type == "Boulders"] +
0.06 * Percentage[Type == "Cobble"] +
0.04 * Percentage[Type == "Gravel"] +
0.03 * Percentage[Type == "Sand"] +
0.02 * Percentage[Type == "Organic_matter"] +
0.01 * Percentage[Type == "Mud"],
na.rm = TRUE
),
Substrate_CV = {
substrate_vals <- Percentage[Type %in% c("Bedrock", "Boulders", "Cobble", "Gravel", "Sand", "Mud", "Organic_matter")]
substrate_vals <- substrate_vals[!is.na(substrate_vals)]
if (length(substrate_vals) > 1 && mean(substrate_vals) > 0) {
sd(substrate_vals) / mean(substrate_vals)
} else {
NA_real_
}
},
.groups = "drop"
) %>%
mutate(
Habitat_Type = case_when(
Rocky_Percentage > 25 ~ "Rocky",
Emergent_Percentage > 25 ~ "Emergent Macrophyte",
Sand_Percentage >= Mud_Percentage ~ "Sandy",
TRUE ~ "Muddy"
)
) %>%
dplyr::select(Monitoring_ID, Habitat_Type, Substrate_index, Substrate_CV)

Monitoring_CPUE_data <- Monitoring_CPUE_data %>%
left_join(habitat_classification, by = "Monitoring_ID")

writexl::write_xlsx(Monitoring_CPUE_data, file.path(der_data_dir, "Monitoring_CPUE_data.xlsx"))
write.csv(Monitoring_CPUE_data, file.path(der_data_dir, "Monitoring_CPUE_data.csv"), row.names = FALSE)
write.csv(habitat_classification, file.path(der_data_dir, "habitat_classification.csv"), row.names = FALSE)

#head(Monitoring_CPUE_data)

```

## habitat-correlation
```{r}
#| label: habitat-correlation

cor(habitat_classification$Substrate_index,
habitat_classification$Substrate_CV,
use = "complete.obs")
```

# Environmental and fish summaries (tables)
```{r}
#| label: env-fish-summary

M_C_data <- Monitoring_CPUE_data
lake_order <- c("Rotorua", "Rotoiti", "Rotoehu", "Rotomā", "Ōkāreka")
M_C_data$Lake <- factor(M_C_data$Lake, levels = lake_order)

ci95 <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 2) return(c(NA_real_, NA_real_))
  se <- sd(x) / sqrt(n)
  tcrit <- qt(0.975, df = n - 1)
  m <- mean(x)
  c(m - tcrit * se, m + tcrit * se)
}

unit_lookup <- Monitoring_data %>%
  dplyr::select(Parameter, Unit) %>%
  dplyr::distinct() %>%
  dplyr::mutate(
    Variable = dplyr::case_when(
      Parameter == "Riparian_vegetation"       ~ "Riparian_vegetation",
      Parameter == "Overhanging_trees"         ~ "Overhanging_trees",
      Parameter == "Wood_cover"                ~ "Wood_cover",
      Parameter == "Temperature"               ~ "Temperature",
      Parameter == "DO_mgl"                    ~ "DO_mgl",
      Parameter == "pH"                        ~ "pH",
      Parameter == "Specific_conductivity"     ~ "Specific_conductivity",
      Parameter == "Substrate_index"           ~ "Substrate_index",
      Parameter == "Slope_5m"                  ~ "Slope_5m",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(Variable)) %>%
  dplyr::select(Variable, Unit) %>%
  dplyr::distinct()

# Add units for derived variables (not in Monitoring_data)
derived_units <- tibble::tibble(
  Variable = c("Emergent_vegetation", "Submerged_vegetation"),
  Unit     = c("%", "%")
)

# Fish presence units (binary)
fish_units <- tibble::tibble(
  Variable = c(
    "Presence_Eel","Presence_Common_smelt","Presence_Catfish",
    "Presence_Goldfish","Presence_Kōaro","Presence_Trout"
  ),
  Unit = ""
)

unit_lookup_all <- dplyr::bind_rows(unit_lookup, derived_units, fish_units) %>%
  dplyr::distinct(Variable, .keep_all = TRUE)

# ---- Environmental summary ----
Env_data <- M_C_data %>%
  dplyr::mutate(
    Emergent_vegetation  = Emergent_Native + Emergent_Non_Native,
    Submerged_vegetation = Submerged_Native + Submerged_Non_Native
  ) %>%
  dplyr::select(
    Lake, Substrate_index, Slope_5m, Riparian_vegetation, Overhanging_trees, Wood_cover,
    Temperature, DO_mgl, pH, Specific_conductivity, Emergent_vegetation, Submerged_vegetation
  ) %>%
  tidyr::pivot_longer(-Lake, names_to = "Variable", values_to = "Value")

Env_summary_table <- Env_data %>%
  dplyr::group_by(Lake, Variable) %>%
  dplyr::summarise(
    n       = sum(!is.na(Value)),
    Mean    = mean(Value, na.rm = TRUE),
    Median  = median(Value, na.rm = TRUE),
    Min     = min(Value, na.rm = TRUE),
    Max     = max(Value, na.rm = TRUE),
    CI_low  = ci95(Value)[1],
    CI_high = ci95(Value)[2],
    .groups = "drop"
  )

Env_summary_all <- Env_data %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    Lake    = "All lakes",
    n       = sum(!is.na(Value)),
    Mean    = mean(Value, na.rm = TRUE),
    Median  = median(Value, na.rm = TRUE),
    Min     = min(Value, na.rm = TRUE),
    Max     = max(Value, na.rm = TRUE),
    CI_low  = ci95(Value)[1],
    CI_high = ci95(Value)[2],
    .groups = "drop"
  ) %>%
  dplyr::select(Lake, dplyr::everything())

Env_summary_table <- dplyr::bind_rows(Env_summary_table, Env_summary_all)

# ---- Fish presence summary ----
Fish_data_long <- M_C_data %>%
  dplyr::select(
    Lake, Presence_Eel, Presence_Common_smelt, Presence_Catfish,
    Presence_Goldfish, Presence_Kōaro, Presence_Trout
  ) %>%
  tidyr::pivot_longer(-Lake, names_to = "Variable", values_to = "Presence")

Fish_summary_table <- Fish_data_long %>%
  dplyr::group_by(Lake, Variable) %>%
  dplyr::summarise(
    n       = sum(!is.na(Presence)),
    Mean    = mean(Presence, na.rm = TRUE),
    Median  = median(Presence, na.rm = TRUE),
    Min     = min(Presence, na.rm = TRUE),
    Max     = max(Presence, na.rm = TRUE),
    CI_low  = ci95(Presence)[1],
    CI_high = ci95(Presence)[2],
    .groups = "drop"
  )

Fish_summary_all <- Fish_data_long %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    Lake    = "All lakes",
    n       = sum(!is.na(Presence)),
    Mean    = mean(Presence, na.rm = TRUE),
    Median  = median(Presence, na.rm = TRUE),
    Min     = min(Presence, na.rm = TRUE),
    Max     = max(Presence, na.rm = TRUE),
    CI_low  = ci95(Presence)[1],
    CI_high = ci95(Presence)[2],
    .groups = "drop"
  ) %>%
  dplyr::select(Lake, dplyr::everything())

Fish_summary_table <- dplyr::bind_rows(Fish_summary_table, Fish_summary_all)

# ---- Combine + attach units + make Variable labels pretty ----
EnvBio_summary_table <- dplyr::bind_rows(Env_summary_table, Fish_summary_table) %>%
  dplyr::left_join(unit_lookup_all, by = "Variable") %>%
  dplyr::mutate(
    Variable = dplyr::recode(
      Variable,
      Substrate_index       = "Substrate index",
      Slope_5m              = "Slope 5m",
      Riparian_vegetation   = "Riparian vegetation",
      Overhanging_trees     = "Overhanging trees",
      Wood_cover            = "Wood cover",
      Emergent_vegetation   = "Emergent vegetation",
      Submerged_vegetation  = "Submerged vegetation",
      Temperature           = "Temperature",
      DO_mgl                = "Dissolved oxygen",
      Specific_conductivity = "Specific conductivity",
      pH                    = "pH",
      Presence_Catfish      = "Presence Catfish",
      Presence_Eel          = "Presence Eel",
      Presence_Goldfish     = "Presence Goldfish",
      Presence_Common_smelt = "Presence Common smelt",
      Presence_Kōaro        = "Presence Kōaro",
      Presence_Trout        = "Presence Trout"
    ),
    Unit = dplyr::coalesce(Unit, "")
  )

lake_order_exact <- c("Rotorua", "Rotoiti", "Rotoehu", "Rotomā", "Ōkāreka", "All lakes")

var_order_exact <- c("Substrate index","Slope 5m","Riparian vegetation","Overhanging trees","Wood cover","Emergent vegetation", "Submerged vegetation", "Temperature", "Dissolved oxygen", "Specific conductivity","pH", "Presence Catfish", "Presence Eel", "Presence Goldfish", "Presence Common smelt", "Presence Kōaro", "Presence Trout")

EnvBio_summary_table <- EnvBio_summary_table %>%
  dplyr::mutate(
    Lake     = factor(as.character(Lake), levels = lake_order_exact),
    Variable = factor(as.character(Variable), levels = var_order_exact)
  ) %>%
  dplyr::select(Lake, Variable, Unit, n, Mean, Median, Min, Max, CI_low, CI_high) %>%
  dplyr::arrange(Variable, Lake) %>%
  dplyr::mutate(
    Lake     = as.character(Lake),
    Variable = as.character(Variable)
  )

# Save if you still want the CSV
write.csv(
  EnvBio_summary_table,
  file = file.path(out_dir, "EnvBio_summary_table.csv"),
  row.names = FALSE
)

knitr::kable(EnvBio_summary_table, caption = "Table S1 Distribution of environmental and biotic variables measured at littoral sampling sites across five Te Arawa lakes in the Rotorua region of Aotearoa New Zealand.",
  digits = 2, align = c("l","l","l","r","r","r","r","r","r","r"))
```

# Kōura response summaries by lake
```{r}
#| label: koura-by-lake
#| fig-width: 3.1
#| fig-height: 4.7
#| dpi: 300
#| fig-cap: "Fig. 2 Kōura presence, CPUE, and BPUE across lakes. The upper panel shows the proportion of sampled sites with kōura present in each lake, with error bars indicating 95% binomial confidence intervals. The middle and lower panels show the distribution of kōura CPUE and BPUE, respectively, across lakes using boxplots (median, interquartile range, and range)."

plot_koura_stats <- function(data, y_var, y_label,
type = c("continuous", "presence"),
show_x_title = FALSE) {

type <- match.arg(type)
lakes <- levels(data$Lake)
xlab_text <- if (show_x_title) "Lake" else NULL

base_theme <- theme_classic() +
theme(
text = element_text(family = "Arial", size = 8),
axis.title = element_text(face = "plain"),
axis.text = element_text(face = "plain"),
panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3)
)

if (type == "continuous") {


ggplot(data, aes(Lake, .data[[y_var]])) +
  geom_boxplot(fill = NA, colour = "black", linewidth = 0.3, outlier.size = 1.5) +
  labs(y = y_label, x = xlab_text) +
  base_theme
} else {
presence_summary <- data %>%
  group_by(Lake) %>%
  summarise(
    n = sum(!is.na(.data[[y_var]])),
    k = sum(.data[[y_var]] == 1, na.rm = TRUE),
    Presence_Rate = ifelse(n > 0, k / n, NA_real_),
    .groups = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    CI_low  = ifelse(n > 0, binom.test(k, n)$conf.int[1], NA_real_),
    CI_high = ifelse(n > 0, binom.test(k, n)$conf.int[2], NA_real_)
  ) %>%
  ungroup() %>%
  mutate(Lake = factor(Lake, levels = lakes))

ggplot(presence_summary, aes(x = Lake, y = Presence_Rate)) +
  geom_col(fill = NA, colour = "black", linewidth = 0.3) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.15, linewidth = 0.3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = xlab_text, y = y_label) 
}
}

KPRES_plot <- plot_koura_stats(M_C_data, "Presence_Kōura", "Kōura presence", type = "presence", show_x_title = FALSE)
KCPUE_plot <- plot_koura_stats(M_C_data, "Weighted_CPUE_Kōura", "Kōura CPUE", type = "continuous", show_x_title = FALSE)
KBPUE_plot <- plot_koura_stats(M_C_data, "Weighted_BPUE_Kōura", "Kōura BPUE", type = "continuous", show_x_title = TRUE)

Koura_plots <- KPRES_plot / KCPUE_plot / KBPUE_plot

ggsave(file.path(out_dir, "Koura_plots.png"), Koura_plots, 
       width = 80, height = 120, units = "mm", dpi = 1200)

Koura_plots
```

# Lake differences glmm
```{r}
#| label: lake_differences_glmm
#| code-fold: true


fit_presence_glmm <- function(data, response, lake_var = "Lake", random_effect = "Habitat_Type") {
  form <- as.formula(paste0(response, " ~ ", lake_var, " + (1|", random_effect, ")"))
  glmmTMB::glmmTMB(form, data = data, family = binomial(link = "logit"))
}

fit_tweedie_glmm <- function(data, response, lake_var = "Lake", random_effect = "Habitat_Type") {
  form <- as.formula(paste0(response, " ~ ", lake_var, " + (1|", random_effect, ")"))
  glmmTMB::glmmTMB(form, data = data, family = glmmTMB::tweedie(link = "log"))
}

m_koura_pres <- fit_presence_glmm(M_C_data, response = "Presence_Kōura",       lake_var = "Lake", random_effect = "Habitat_Type")
m_koura_cpue <- fit_tweedie_glmm( M_C_data, response = "Weighted_CPUE_Kōura",  lake_var = "Lake", random_effect = "Habitat_Type")
m_koura_bpue <- fit_tweedie_glmm( M_C_data, response = "Weighted_BPUE_Kōura",  lake_var = "Lake", random_effect = "Habitat_Type")

# Overall lake effect (LRT via drop1)
pres_lrt <- drop1(m_koura_pres, test = "Chisq")
cpue_lrt <- drop1(m_koura_cpue, test = "Chisq")
bpue_lrt <- drop1(m_koura_bpue, test = "Chisq")

# Pairwise lake comparisons (BH-adjusted)
pres_pairs <- pairs(emmeans(m_koura_pres, ~ Lake, type = "response"), adjust = "BH")
cpue_pairs <- pairs(emmeans(m_koura_cpue, ~ Lake, type = "response"), adjust = "BH")
bpue_pairs <- pairs(emmeans(m_koura_bpue, ~ Lake, type = "response"), adjust = "BH")

pres_lrt
cpue_lrt
bpue_lrt

```

# GAM modelling
## Helpers
```{r}
#| label: gam-helpers

Modeling_data <- Monitoring_CPUE_data

alpha_sig    <- 0.1
p_cutoff_ml  <- 0.05
vif_thresh   <- 5
INCLUDE_RE   <- TRUE

custom_k <- list(
Slope_5m             = 10,
Riparian_vegetation  = 7,
Overhanging_trees    = 5,
Wood_cover           = 10,
Substrate_index      = 10,
Temperature          = 10,
pH                   = 10,
DO_mgl               = 10,
Emergent_Native      = 9,
Submerged_Non_Native = 9,
Turf_Native          = 6,
Submerged_Native     = 6
)

fish_vars <- c(
"Presence_Goldfish",
"Presence_Eel",
"Presence_Catfish",
"Presence_Common_smelt",
"Presence_Kōaro"
)

vars_common <- c(
"LID",
"Slope_5m","Riparian_vegetation","Overhanging_trees","Wood_cover","Substrate_index",
"Temperature","pH","DO_mgl","DO_percent","Specific_conductivity",
"Emergent_Native","Submerged_Non_Native","Turf_Native","Submerged_Native","Emergent_Non_Native",
fish_vars
)

is_cont <- function(x) is.numeric(x) && dplyr::n_distinct(x, na.rm = TRUE) >= 5

build_smooth_gam <- function(var, data, custom_k = list(), include_re_for_LID = FALSE) {
if (identical(var, "LID") && include_re_for_LID) return("s(LID, bs='re')")
x <- data[[var]]
if (is.numeric(x)) {
nuniq <- dplyr::n_distinct(x, na.rm = TRUE)
if (nuniq < 5) return(var)
k_req <- if (!is.null(custom_k[[var]])) custom_k[[var]] else 10
k_cap <- max(3, min(k_req, nuniq - 1))
return(paste0("s(", var, ", bs='ts', k=", k_cap, ")"))
}
var
}

exclude_if_RE <- function(mod){
if (!length(mod$smooth)) return(NULL)
has_re <- vapply(mod$smooth, function(s) "LID" %in% s$term, logical(1))
if (any(has_re)) "s(LID)" else NULL
}

remove_high_vif_glmBI <- function(
data,
response,
predictors,
threshold    = vif_thresh,
protect_vars = character(0),
verbose      = FALSE
) {
nzv <- caret::nearZeroVar(data[, predictors, drop = FALSE])
if (length(nzv)) predictors <- predictors[-nzv]

rpt <- list(removed = character(), start = predictors)
last_vif <- NULL

repeat {
if (!length(predictors)) stop("All predictors removed during VIF pruning.")


fml <- as.formula(paste(response, "~", paste(predictors, collapse = " + ")))
model <- try(lm(fml, data = data), silent = TRUE)
if (inherits(model, "try-error")) break

vif_data <- performance::check_collinearity(model)
vif_data <- vif_data[!grepl("\\|", vif_data$Term), , drop = FALSE]
last_vif <- vif_data

if (!nrow(vif_data) || all(vif_data$VIF < threshold)) break

ord <- order(vif_data$VIF, decreasing = TRUE)
to_remove <- NA_character_
for (i in ord) {
  cand <- vif_data$Term[i]
  if (!(cand %in% protect_vars)) { to_remove <- cand; break }
}
if (is.na(to_remove)) break

predictors  <- setdiff(predictors, to_remove)
rpt$removed <- c(rpt$removed, to_remove)
}

rpt$final_vif <- last_vif
list(predictors = predictors, report = rpt)
}

remove_high_vif_glmmTMB <- function(
data,
response,
predictors,
threshold    = vif_thresh,
protect_vars = character(0),
verbose      = FALSE
) {
nzv <- caret::nearZeroVar(data[, predictors, drop = FALSE])
if (length(nzv)) predictors <- predictors[-nzv]

rpt <- list(removed = character(), start = predictors)
last_vif <- NULL

repeat {
if (!length(predictors)) stop("All predictors removed during VIF pruning.")


fml <- as.formula(paste(response, "~", paste(predictors, collapse = " + ")))
model <- try(lm(fml, data = data), silent = TRUE)
if (inherits(model, "try-error")) break

vif_data <- performance::check_collinearity(model)
vif_data <- vif_data[!grepl("\\|", vif_data$Term), , drop = FALSE]
last_vif <- vif_data

if (!nrow(vif_data) || all(vif_data$VIF < threshold)) break

ord <- order(vif_data$VIF, decreasing = TRUE)
to_remove <- NA_character_
for (i in ord) {
  cand <- vif_data$Term[i]
  if (!(cand %in% protect_vars)) { to_remove <- cand; break }
}
if (is.na(to_remove)) break

predictors  <- setdiff(predictors, to_remove)
rpt$removed <- c(rpt$removed, to_remove)

}

rpt$final_vif <- last_vif
list(predictors = predictors, report = rpt)
}

as_plot <- function(p) if (inherits(p, c("gg","ggplot","patchwork"))) p else patchwork::plot_spacer()

prepare_block <- function(Modeling_data, response, vars_common, id = "LID"){
vars <- c(response, vars_common)
Modeling_data %>%
dplyr::select(dplyr::all_of(vars)) %>%
dplyr::mutate(
"{id}" := factor(.data[[id]]),
dplyr::across(dplyr::any_of(fish_vars), ~ as.numeric(.x))
)
}

ref_row <- function(data){
dplyr::summarise(
data,
dplyr::across(
dplyr::everything(),
\(x){
if (is.numeric(x)) stats::median(x, na.rm = TRUE)
else if (is.factor(x)) levels(x)[1L]
else if (is.logical(x)) FALSE
else if (is.character(x)) unique(stats::na.omit(x))[1L]
else x[1L]
}
)
)
}

is_binary_numeric <- function(x) is.numeric(x) && dplyr::n_distinct(x, na.rm = TRUE) == 2

get_term_p <- function(m, var, kind = c("any","param","smooth")){
kind <- match.arg(kind)
sm <- summary(m)

if (kind %in% c("any","param")) {
if (!is.null(sm$p.table) && nrow(sm$p.table) > 0) {
pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)","Pr(>|z|)"))[1]
if (!is.na(pcol) && var %in% rownames(sm$p.table)) {
return(as.numeric(sm$p.table[var, pcol]))
}
}
}

if (kind %in% c("any","smooth")) {
if (!is.null(sm$s.table) && nrow(sm$s.table) > 0) {
sname <- paste0("s(", var, ")")
if (sname %in% rownames(sm$s.table)) {
return(as.numeric(sm$s.table[sname, "p-value"]))
}
}
}

NA_real_
}

plot_binary_single_component <- function(model, var, data,
family = c("binomial","gamma"),
S = 2000, alpha = 0.05,
exclude_RE = TRUE,
hold_binaries = c("mean","zero","one"),
id = "LID",
fish_covars = c("Presence_Goldfish","Presence_Eel","Presence_Catfish","Presence_Common_smelt"),
seed = 1) {
family <- match.arg(family)
hold_binaries <- match.arg(hold_binaries)

tl <- attr(terms(model), "term.labels")
has_term <- any(tl == var) || any(grepl(paste0("^s\\(", var, "(,|\\))"), tl))
if (!has_term) return(ggplot() + theme_void())

num_means <- data %>%
dplyr::summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
base <- as.list(num_means)
if (id %in% names(data) && is.factor(data[[id]]))
base[[id]] <- factor(levels(data[[id]])[1], levels = levels(data[[id]]))

others <- setdiff(intersect(fish_covars, names(data)), var)
set_bin <- function(x){
if (hold_binaries == "mean") round(mean(data[[x]], na.rm = TRUE))
else if (hold_binaries == "zero") 0L else 1L
}
for (x in others) base[[x]] <- set_bin(x)

mk_nd <- function(level){
nd <- as.data.frame(base, stringsAsFactors = FALSE)
nd[[var]] <- as.integer(level)
if (id %in% names(data) && is.factor(data[[id]])) {
nd <- do.call(
rbind,
lapply(levels(data[[id]]), function(lv){
r <- nd
r[[id]] <- factor(lv, levels = levels(data[[id]]))
r
})
)
}
if (!is.null(model$model)) {
common <- intersect(names(nd), names(model$model))
for (nm in common) if (is.factor(model$model[[nm]]))
nd[[nm]] <- factor(nd[[nm]], levels = levels(model$model[[nm]]))
}
nd
}
nd0 <- mk_nd(0L)
nd1 <- mk_nd(1L)
nd  <- dplyr::bind_rows(
dplyr::mutate(nd0, .level = 0L),
dplyr::mutate(nd1, .level = 1L)
)

excl <- if (exclude_RE) exclude_if_RE(model) else NULL
lp <- predict(model, newdata = nd, type = "link", se.fit = TRUE, exclude = excl)

set.seed(seed)
n <- nrow(nd)
Z <- matrix(
rnorm(n * S, lp$fit, pmax(lp$se.fit, .Machine$double.eps)),
nrow = n, ncol = S
)

if (family == "binomial") {
Y <- plogis(Z); ylab <- "Presence probability"
} else {
Y <- exp(Z);    ylab <- "Mean given presence (μ)"
}

sim_by_level <- function(level_flag){
sims <- Y[nd$.level == level_flag, , drop = FALSE]
colMeans(sims)
}
sim0 <- sim_by_level(0L)
sim1 <- sim_by_level(1L)

qlo <- alpha/2
qhi <- 1 - alpha/2
df <- data.frame(
level = factor(c(0,1), levels = c(0,1), labels = c("Absent","Present")),
mean  = c(mean(sim0), mean(sim1)),
lwr   = c(quantile(sim0, qlo), quantile(sim1, qlo)),
upr   = c(quantile(sim0, qhi), quantile(sim1, qhi))
)

ggplot(df, aes(x = level, y = mean)) +
geom_col(width = 0.6) +
geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2) +
labs(x = var, y = ylab)
}

plot_linear_single_component <- function(model, var, data,
family = c("binomial","gamma"),
n = 100, alpha = 0.05,
exclude_RE = TRUE) {
family <- match.arg(family)

num_means <- data %>%
dplyr::summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
base <- as.list(num_means)

if (!is.null(model$model)) {
for (nm in names(model$model)) {
if (is.factor(model$model[[nm]])) {
base[[nm]] <- factor(
levels(model$model[[nm]])[1],
levels = levels(model$model[[nm]])
)
}
}
}

x  <- data[[var]]
xr <- range(x, na.rm = TRUE)
grid <- seq(xr[1], xr[2], length.out = n)

nd <- as.data.frame(base)
nd <- nd[rep(1, n), , drop = FALSE]
nd[[var]] <- grid

excl <- if (exclude_RE) exclude_if_RE(model) else NULL
pr <- predict(model, newdata = nd, type = "link", se.fit = TRUE, exclude = excl)

if (family == "binomial") {
fit <- plogis(pr$fit)
lwr <- plogis(pr$fit - qnorm(1 - alpha/2) * pr$se.fit)
upr <- plogis(pr$fit + qnorm(1 - alpha/2) * pr$se.fit)
ylab <- "Response (probability)"
} else {
fit <- exp(pr$fit)
lwr <- exp(pr$fit - qnorm(1 - alpha/2) * pr$se.fit)
upr <- exp(pr$fit + qnorm(1 - alpha/2) * pr$se.fit)
ylab <- "Response (μ)"
}

df <- data.frame(x = grid, fit = fit, lwr = lwr, upr = upr)
ggplot(df, aes(x, fit)) +
geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
geom_line() +
labs(x = var, y = ylab)
}

parametric_panel <- function(model, data, alpha = 0.1, exclude = character(0),
family = c("binomial","gamma"),
fish_covars = c("Presence_Goldfish","Presence_Eel","Presence_Catfish","Presence_Common_smelt"),
hold_binaries = "mean") {
family <- match.arg(family)

sm <- summary(model)
if (is.null(sm$p.table) || nrow(sm$p.table) == 0) return(ggplot() + theme_void())

pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)","Pr(>|z|)"))[1]
if (is.na(pcol)) return(ggplot() + theme_void())

par_names <- setdiff(rownames(sm$p.table), "(Intercept)")

smooth_rows <- if (!is.null(sm$s.table)) rownames(sm$s.table) else character(0)
smooth_vars <- sub("^s\\(([^,\\)]+).*$", "\\1", smooth_rows)
par_names <- setdiff(par_names, union(exclude, smooth_vars))

par_sig <- par_names[sm$p.table[par_names, pcol, drop = TRUE] <= alpha]
if (!length(par_sig)) return(ggplot() + theme_void())

plots <- lapply(par_sig, function(v) {
x <- data[[v]]
if (is.null(x)) return(ggplot() + theme_void())

is_bin_num <- is.numeric(x) && dplyr::n_distinct(x, na.rm = TRUE) <= 2
is_bin_fac <- is.factor(x)  && nlevels(x) == 2

if (is_bin_num || is_bin_fac) {
plot_binary_single_component(
model, v, data,
family        = family,
hold_binaries = hold_binaries,
fish_covars   = fish_covars
)
} else {
plot_linear_single_component(model, v, data, family = family)
}
})

if (!length(plots)) return(ggplot() + theme_void())
patchwork::wrap_plots(plots, ncol = 1)
}

```

## Occupancy model (presence/absence)
```{r}
#| label: occupancy-model
#| fig-width: 7.1
#| fig-height: 4.7
#| dpi: 300
#| results: hide
#| warning: false
#| message: false
#| fig-cap: "Fig. 3 Kōura occupancy GAMM results and model performance. a) Estimated smooth terms for riparian vegetation, substrate index, surface water temperature, and specific conductivity, shown as partial effects with 95% confidence intervals. b) Receiver operating characteristic (ROC) curve illustrating model discrimination, with the area under the curve (AUC) reported. c) Calibration plot based on five equal-frequency bins, comparing mean predicted occupancy probabilities with observed proportions of occupied sites; the dashed 1:1 line indicates perfect calibration."


PData <- prepare_block(Modeling_data, "Presence_Kōura", vars_common)

id <- "LID"
pred_fixed_occ <- setdiff(names(PData), c("Presence_Kōura", id))

vif_occ <- remove_high_vif_glmBI(
PData,
"Presence_Kōura",
pred_fixed_occ,
threshold    = vif_thresh,
protect_vars = character(0)
)
kept_fixed_occ <- vif_occ$predictors

vars_step_occ <- c(kept_fixed_occ, if (INCLUDE_RE) id)
rhs_full_occ <- paste(
vapply(
vars_step_occ,
function(v) build_smooth_gam(v, PData, custom_k, include_re_for_LID = INCLUDE_RE),
character(1)
),
collapse = " + "
)
form_full_occ <- as.formula(paste("Presence_Kōura ~", rhs_full_occ))

full_ml_occ <- mgcv::gam(
form_full_occ,
data   = PData,
family = binomial(link = "logit"),
method = "ML",
select = TRUE
)

remaining_occ <- vars_step_occ
protected_step_vars <- if (INCLUDE_RE) id else character(0)

repeat {
rhs_now <- paste(
vapply(
remaining_occ,
function(v) build_smooth_gam(v, PData, custom_k, include_re_for_LID = INCLUDE_RE),
character(1)
),
collapse = " + "
)
m_now <- mgcv::gam(
as.formula(paste("Presence_Kōura ~", rhs_now)),
data   = PData,
family = binomial(link = "logit"),
method = "ML",
select = TRUE
)
sm <- summary(m_now)

ps <- c()
if (!is.null(sm$p.table) && nrow(sm$p.table) > 0) {
pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)","Pr(>|z|)"))[1]
pvec <- sm$p.table[, pcol]
ps <- c(ps, pvec[names(pvec) != "(Intercept)"])
}
if (!is.null(sm$s.table) && nrow(sm$s.table) > 0) {
pvec <- sm$s.table[, "p-value"]
names(pvec) <- rownames(sm$s.table)
ps <- c(ps, pvec)
}
drop_candidates <- ps[ps > p_cutoff_ml]
if (!length(drop_candidates)) { red_ml_occ <- m_now; break }

ordered <- names(sort(drop_candidates, decreasing = TRUE))
ordered_vars <- vapply(
ordered,
function(x) if (grepl("^s\\(", x)) sub("^s\\(([^,]+).*\\)$", "\\1", x) else x,
character(1)
)

ordered_vars <- setdiff(ordered_vars, protected_step_vars)
if (!length(ordered_vars)) { red_ml_occ <- m_now; break }

remove_v <- ordered_vars[1]
remaining_occ <- setdiff(remaining_occ, remove_v)
if (!length(remaining_occ)) { red_ml_occ <- m_now; break }
}

final_occ <- mgcv::gam(
formula(red_ml_occ),
data   = PData,
family = binomial(link = "logit"),
method = "REML",
select = TRUE
)

# Smooth plots: keep draw() outputs as-is (they may be ggplot OR patchwork)
sm <- gratia::smooth_estimates(final_occ)
smooth_names <- unique(sm$.smooth)
smooth_4_names <- smooth_names[1:min(4, length(smooth_names))]

plots_4 <- lapply(smooth_4_names, function(s) {
  p <- gratia::draw(final_occ, select = s, scales = "fixed", se = FALSE) +
    labs(title = clean_smooth_title(s), x = NULL, y = "Partial effect") +
    base_theme_bw
  p
})

# Combine without forcing wrap_dims assumptions about panel counts
final_p_occ_smooths <- patchwork::wrap_plots(plots_4, ncol = length(plots_4))

# Parametric panel (unchanged call)
p_occ_param <- parametric_panel(
  final_occ,
  PData,
  alpha       = 0.1,
  exclude     = c("LID"),
  family      = "binomial",
  fish_covars = fish_vars
) + base_theme_bw

# ROC on fitted model predictions
pred_pres <- predict(final_occ, type = "response", exclude = exclude_if_RE(final_occ))
obs_pres  <- PData$Presence_Kōura == 1
roc_obj   <- pROC::roc(obs_pres, pred_pres)
auc_val   <- as.numeric(pROC::auc(roc_obj))

roc_df <- data.frame(
  tpr = roc_obj$sensitivities,
  fpr = 1 - roc_obj$specificities
)

p_pres_roc <- ggplot(roc_df, aes(fpr, tpr)) +
  geom_path(colour = "black", linewidth = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black", linewidth = 0.4) +
  annotate("text", x = 0.65, y = 0.1, label = paste0("AUC = ", round(auc_val, 3)), size = 3) +
  labs(x = "False positive rate", y = "True positive rate") +
  base_theme_bw

# ---- CV by LID (fix unseen LID warnings by forcing LID in test to a known training level) ----
set.seed(1)
k_occ      <- min(5, nlevels(PData$LID))
lid_levels <- levels(PData$LID)
lid_folds  <- sample(rep(1:k_occ, length.out = length(lid_levels)))
lid2fold   <- setNames(lid_folds, lid_levels)
fold_vec   <- unname(lid2fold[as.character(PData$LID)])
pred_cv_occ <- rep(NA_real_, nrow(PData))

final_terms <- {
  tl <- attr(terms(final_occ), "term.labels")
  sub("^s\\(([^,\\)]+).*$", "\\1", tl)
}

for (fold in 1:k_occ) {
  idx   <- which(fold_vec == fold)
  train <- PData[-idx, , drop = FALSE]
  test  <- PData[idx,  , drop = FALSE]

  rhs_now <- paste(
    vapply(
      final_terms,
      function(v) build_smooth_gam(v, train, custom_k, include_re_for_LID = INCLUDE_RE),
      character(1)
    ),
    collapse = " + "
  )
  fml_now <- as.formula(paste("Presence_Kōura ~", rhs_now))

  m <- mgcv::gam(
    fml_now,
    data   = train,
    family = binomial(link = "logit"),
    method = "REML",
    select = TRUE
  )

  # Align factor levels; then neutralise LID for prediction if RE is excluded
  for (nm in names(test)) {
    if (is.factor(train[[nm]])) {
      test[[nm]] <- factor(test[[nm]], levels = levels(train[[nm]]))
    }
  }

  if ("LID" %in% names(test) && is.factor(test$LID)) {
    test$LID <- factor(levels(train$LID)[1L], levels = levels(train$LID))
  }

  pred_cv_occ[idx] <- predict(m, newdata = test, type = "response", exclude = exclude_if_RE(m))
}

calib_data_occ <- tibble::tibble(pred = pred_cv_occ, obs = PData$Presence_Kōura) %>%
  dplyr::mutate(bin = dplyr::ntile(pred, 5)) %>%
  dplyr::group_by(bin) %>%
  dplyr::summarise(
    mean_pred = mean(pred, na.rm = TRUE),
    obs_rate  = mean(obs,  na.rm = TRUE),
    n         = dplyr::n(),
    .groups   = "drop"
  )


Calibration_plot_occ <- ggplot(calib_data_occ, aes(mean_pred, obs_rate)) +
  geom_point(size = 2.2, colour = "black") +
  geom_line(colour = "black", linewidth = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black", linewidth = 0.4) +
  labs(x = "Mean predicted probability", y = "Observed proportion") +
  base_theme_bw

# Layout (avoid forcing a layout that conflicts with nested patchworks)
top_row    <- final_p_occ_smooths
bottom_row <- (p_pres_roc + Calibration_plot_occ) + patchwork::plot_layout(ncol = 2)

final_plot_occupancy <- (top_row / bottom_row)+
  patchwork::plot_layout(heights = c(1, 2))

ggsave(filename = file.path(out_dir, "Occupancy_panels.png"), plot = final_plot_occupancy,  width = 180, height = 120, units = "mm", dpi = 1200)

final_plot_occupancy

gam.check(final_occ)
concurvity(final_occ, full = FALSE)
```

## CPUE hurdle model
```{r}
#| label: cpue-hurdle
#| fig-width: 5.1
#| fig-height: 4.7
#| dpi: 300
#| results: hide
#| warning: false
#| message: false
#| fig-cap: "Fig. 4 Kōura CPUE GAMM results and model performance. a) Estimated smooth terms for surface water temperature, pH, and emergent native vegetation, shown with partial effects and 95% confidence intervals. See Fig. S2 for raw data underlying modelled relationships. b) Predicted versus observed CPUE from the positive (Gamma) component of the hurdle model, with predictions excluding the random lake effect. The dashed 1:1 line indicates perfect agreement; annotated values report R² and RMSE to summarise model fit."

CData <- prepare_block(Modeling_data, "Weighted_CPUE_Kōura", vars_common) %>%
dplyr::mutate(CPUE_pos = as.integer(Weighted_CPUE_Kōura > 0))

id <- "LID"

pred_fixed_cpue_pres <- setdiff(
names(CData),
c("CPUE_pos", id, "Weighted_CPUE_Kōura", "Weighted_CPUE_Koura", "CPUE")
)



Cpos <- dplyr::filter(CData, CPUE_pos == 1L)

pred_fixed_cpue_pos <- setdiff(names(Cpos), c("Weighted_CPUE_Kōura", id))

vif_cpue_pos <- remove_high_vif_glmmTMB(
Cpos,
"Weighted_CPUE_Kōura",
pred_fixed_cpue_pos,
threshold    = vif_thresh,
protect_vars = character(0)
)
kept_fixed_cpue_pos <- vif_cpue_pos$predictors

vars_step_cpue_pos <- c(kept_fixed_cpue_pos, if (INCLUDE_RE) id)

rhs_cpue_pos <- paste(
vapply(
vars_step_cpue_pos,
function(v) build_smooth_gam(v, Cpos, custom_k, include_re_for_LID = INCLUDE_RE),
character(1)
),
collapse = " + "
)

full_ml_cpue_pos <- mgcv::gam(
as.formula(paste("Weighted_CPUE_Kōura ~", rhs_cpue_pos)),
data   = Cpos,
family = Gamma(link = "log"),
method = "ML",
select = TRUE
)

remaining_cpue_pos <- vars_step_cpue_pos
protected_step_vars_pos <- if (INCLUDE_RE) id else character(0)

repeat {
rhs_now <- paste(
vapply(
remaining_cpue_pos,
function(v) build_smooth_gam(v, Cpos, custom_k, include_re_for_LID = INCLUDE_RE),
character(1)
),
collapse = " + "
)
m_now <- mgcv::gam(
as.formula(paste("Weighted_CPUE_Kōura ~", rhs_now)),
data   = Cpos,
family = Gamma(link = "log"),
method = "ML",
select = TRUE
)
sm <- summary(m_now)

ps <- c()
if (!is.null(sm$p.table) && nrow(sm$p.table) > 0) {
pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)", "Pr(>|z|)"))[1]
pvec <- sm$p.table[, pcol]
ps   <- c(ps, pvec[names(pvec) != "(Intercept)"])
}
if (!is.null(sm$s.table) && nrow(sm$s.table) > 0) {
pvec <- sm$s.table[, "p-value"]
names(pvec) <- rownames(sm$s.table)
ps   <- c(ps, pvec)
}
drop_candidates <- ps[ps > p_cutoff_ml]
if (!length(drop_candidates)) { red_ml_cpue_pos <- m_now; break }

ordered <- names(sort(drop_candidates, decreasing = TRUE))
ordered_vars <- vapply(
ordered,
function(x) if (grepl("^s\\(", x)) sub("^s\\(([^,]+).*\\)$", "\\1", x) else x,
character(1)
)

ordered_vars <- setdiff(ordered_vars, protected_step_vars_pos)
if (!length(ordered_vars)) { red_ml_cpue_pos <- m_now; break }

remove_v <- ordered_vars[1]
remaining_cpue_pos <- setdiff(remaining_cpue_pos, remove_v)
if (!length(remaining_cpue_pos)) { red_ml_cpue_pos <- m_now; break }
}

final_cpue_pos <- mgcv::gam(
formula(red_ml_cpue_pos),
data   = Cpos,
family = Gamma(link = "log"),
method = "REML",
select = TRUE
)

sm_cpue <- smooth_estimates(final_cpue_pos)
cpue_smooth_names <- setdiff(unique(sm_cpue$.smooth), "s(LID)")
cpue_smooth_plots <- lapply(cpue_smooth_names, function(s) {
gratia::draw(final_cpue_pos, select = s, scales = "fixed", se = FALSE) +
labs(title = clean_smooth_title(s), x = NULL, y = "Partial effect") +
base_theme_bw
})

p_cpue_pos_smooths <- patchwork::wrap_plots(cpue_smooth_plots, ncol = length(cpue_smooth_plots)) +
patchwork::plot_annotation(tag_levels = "a")

pred_pos <- predict(final_cpue_pos, type = "response")
obs_pos  <- final_cpue_pos$y
R2   <- cor(pred_pos, obs_pos, use = "complete.obs")^2
RMSE <- sqrt(mean((pred_pos - obs_pos)^2, na.rm = TRUE))

df_scat <- data.frame(Predicted = pred_pos, Observed = obs_pos)

p_cpue_pos_scatter <- ggplot(df_scat, aes(Predicted, Observed)) +
geom_point(size = 1.8, colour = "black") +
geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black", linewidth = 0.4) +
annotate(
"text",
x = max(df_scat$Predicted, na.rm = TRUE) * 0.7,
y = max(df_scat$Observed,  na.rm = TRUE) * 0.8,
label = paste0("R\u00B2 = ", round(R2, 2), "\nRMSE = ", round(RMSE, 2)),
size = 3
) +
labs(x = "Predicted CPUE", y = "Observed CPUE", tag = "b") +
base_theme_bw

CPUE_panel_positives <- (p_cpue_pos_smooths / p_cpue_pos_scatter) +
patchwork::plot_layout(heights = c(1, 2))

ggsave(filename = file.path(out_dir, "CPUE_panel_positives.png"),plot = CPUE_panel_positives,width = 130, height = 120, units = "mm", dpi = 1200)

CPUE_panel_positives

gam.check(final_cpue_pos) 
concurvity(final_cpue_pos, full = FALSE)

```

## BPUE hurdle model
```{r}
#| label: bpue-hurdle
#| fig-width: 3.9
#| fig-height: 4.7
#| dpi: 300
#| results: hide
#| warning: false
#| message: false
#| fig-cap: "Fig. 5 Kōura BPUE GAMM results and model performance. a) Estimated smooth terms for pH and emergent native vegetation, shown with partial effects and 95% confidence intervals. See Fig. S2 for raw data underlying modelled relationships. b) Predicted versus observed BPUE from the positive (Gamma) component of the hurdle model, with predictions excluding the random lake effect. The dashed 1:1 line indicates perfect agreement; annotated values report R² and RMSE to summarise model fit."

BData <- prepare_block(Modeling_data, "Weighted_BPUE_Kōura", vars_common) %>%
dplyr::mutate(BPUE_pos = as.integer(Weighted_BPUE_Kōura > 0))

id <- "LID"

pred_fixed_BPUE_pres <- setdiff(names(BData),c("BPUE_pos", id, "Weighted_BPUE_Kōura", "Weighted_BPUE_Koura", "Weighted_BPUE"))

Bpos <- dplyr::filter(BData, BPUE_pos == 1L)

pred_fixed_BPUE_pos <- setdiff(names(Bpos), c("Weighted_BPUE_Kōura", id))

vif_BPUE_pos <- remove_high_vif_glmmTMB(Bpos,"Weighted_BPUE_Kōura",pred_fixed_BPUE_pos,threshold = vif_thresh,protect_vars = character(0))
kept_fixed_BPUE_pos <- vif_BPUE_pos$predictors

vars_step_BPUE_pos <- c(kept_fixed_BPUE_pos, if (INCLUDE_RE) id)

rhs_BPUE_pos <- paste(vapply(vars_step_BPUE_pos,function(v) build_smooth_gam(v, Bpos, custom_k, include_re_for_LID = INCLUDE_RE),
character(1)),
collapse = " + "
)

full_ml_BPUE_pos <- mgcv::gam(
as.formula(paste("Weighted_BPUE_Kōura ~", rhs_BPUE_pos)),
data   = Bpos,
family = Gamma(link = "log"),
method = "ML",
select = TRUE
)

remaining_BPUE_pos <- vars_step_BPUE_pos
protected_step_vars_bpos <- if (INCLUDE_RE) id else character(0)

repeat {
rhs_now <- paste(
vapply(
remaining_BPUE_pos,
function(v) build_smooth_gam(v, Bpos, custom_k, include_re_for_LID = INCLUDE_RE),
character(1)
),
collapse = " + "
)
m_now <- mgcv::gam(
as.formula(paste("Weighted_BPUE_Kōura ~", rhs_now)),
data   = Bpos,
family = Gamma(link = "log"),
method = "ML",
select = TRUE
)
sm <- summary(m_now)

ps <- c()
if (!is.null(sm$p.table) && nrow(sm$p.table) > 0) {
pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)", "Pr(>|z|)"))[1]
pvec <- sm$p.table[, pcol]
ps   <- c(ps, pvec[names(pvec) != "(Intercept)"])
}
if (!is.null(sm$s.table) && nrow(sm$s.table) > 0) {
pvec <- sm$s.table[, "p-value"]
names(pvec) <- rownames(sm$s.table)
ps   <- c(ps, pvec)
}
drop_candidates <- ps[ps > p_cutoff_ml]
if (!length(drop_candidates)) { red_ml_BPUE_pos <- m_now; break }

ordered <- names(sort(drop_candidates, decreasing = TRUE))
ordered_vars <- vapply(
ordered,
function(x) if (grepl("^s\\(", x)) sub("^s\\(([^,]+).*\\)$", "\\1", x) else x,
character(1)
)

ordered_vars <- setdiff(ordered_vars, protected_step_vars_bpos)
if (!length(ordered_vars)) { red_ml_BPUE_pos <- m_now; break }

remove_v <- ordered_vars[1]
remaining_BPUE_pos <- setdiff(remaining_BPUE_pos, remove_v)
if (!length(remaining_BPUE_pos)) { red_ml_BPUE_pos <- m_now; break }
}

final_BPUE_pos <- mgcv::gam(
formula(red_ml_BPUE_pos),
data   = Bpos,
family = Gamma(link = "log"),
method = "REML",
select = TRUE
)

sm_bpue <- smooth_estimates(final_BPUE_pos)
bpue_smooth_names <- setdiff(unique(sm_bpue$.smooth), "s(LID)")
bpue_smooth_plots <- lapply(bpue_smooth_names, function(s) {
gratia::draw(final_BPUE_pos, select = s, scales = "fixed", se = FALSE) +
labs(title = clean_smooth_title(s), x = NULL, y = "Partial effect") +
base_theme_bw
})

p_BPUE_pos_smooths <- patchwork::wrap_plots(bpue_smooth_plots, ncol = length(bpue_smooth_plots)) +
patchwork::plot_annotation(tag_levels = "a")

pred_pos_b <- predict(final_BPUE_pos, type = "response")
obs_pos_b  <- final_BPUE_pos$y
R2_b   <- cor(pred_pos_b, obs_pos_b, use = "complete.obs")^2
RMSE_b <- sqrt(mean((pred_pos_b - obs_pos_b)^2, na.rm = TRUE))

df_scat_b <- data.frame(Predicted = pred_pos_b, Observed = obs_pos_b)

p_BPUE_pos_scatter <- ggplot(df_scat_b, aes(Predicted, Observed)) +
geom_point(size = 1.8, colour = "black") +
geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black", linewidth = 0.4) +
annotate(
"text",
x = max(df_scat_b$Predicted, na.rm = TRUE) * 0.7,
y = max(df_scat_b$Observed,  na.rm = TRUE) * 0.8,
label = paste0("R\u00B2 = ", round(R2_b, 2), "\nRMSE = ", round(RMSE_b, 2)),
size = 3
) +
labs(x = "Predicted BPUE", y = "Observed BPUE", tag = "b") +
base_theme_bw

BPUE_panel_positives <- (p_BPUE_pos_smooths / p_BPUE_pos_scatter) +
patchwork::plot_layout(heights = c(1, 2))

ggsave(filename = file.path(out_dir, "BPUE_panel_positives.png"),plot = BPUE_panel_positives,width = 100, height = 120, units = "mm", dpi = 1200)

BPUE_panel_positives

gam.check(final_BPUE_pos)
concurvity(final_BPUE_pos, full = FALSE)

```

## Raw-data-plots CPUE & BPUE
```{r}
#| label: raw-data-plots
#| fig-width: 7
#| fig-height: 4
#| dpi: 300
#| fig-cap: "Fig. S2 Raw relationships between significant habitat predictors and kōura CPUE (upper panels) and BPUE (lower panels) at sites where kōura were present. Points represent observed values at individual sites. Temperature (°C), pH, and emergent native vegetation cover (%) are shown for CPUE; pH and emergent native vegetation cover (%) are shown for BPUE."

make_raw_plot <- function(data, x_var, y_var, y_label) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    geom_point(size = 1.8, colour = "black") +
    labs(x = gsub("_", " ", x_var), y = y_label) +
    base_theme_bw
}

# CPUE raw plots
p_raw_temp_cpue  <- make_raw_plot(Cpos, "Temperature",     "Weighted_CPUE_Kōura", "CPUE")
p_raw_ph_cpue    <- make_raw_plot(Cpos, "pH",              "Weighted_CPUE_Kōura", "CPUE")
p_raw_emerg_cpue <- make_raw_plot(Cpos, "Emergent_Native", "Weighted_CPUE_Kōura", "CPUE")

# BPUE raw plots
p_raw_ph_bpue    <- make_raw_plot(Bpos, "pH",              "Weighted_BPUE_Kōura", "BPUE")
p_raw_emerg_bpue <- make_raw_plot(Bpos, "Emergent_Native", "Weighted_BPUE_Kōura", "BPUE")

raw_plots <- (p_raw_temp_cpue + p_raw_ph_cpue + p_raw_emerg_cpue) /
             (plot_spacer() + p_raw_ph_bpue + p_raw_emerg_bpue)

ggsave(file.path(out_dir, "Raw_data_plots.png"), raw_plots,
       width = 180, height = 120, units = "mm", dpi = 1200)

raw_plots
```

## Model tables (exports)
```{r}
#| label: model-tables

gam_results_table <- function(model, model_name = deparse(substitute(model)), digits = 3) {
sm <- summary(model)

ptab <- as.data.frame(sm$p.table)
ptab <- rownames_to_column(ptab, "term")
ptab <- as_tibble(ptab)

p_col <- names(ptab)[grepl("^Pr\\(>\\|", names(ptab))]
if (length(p_col) != 1) stop("Could not uniquely identify p-value column in p.table.")

stat_col <- names(ptab)[grepl("value$", names(ptab))]
if (length(stat_col) != 1) stop("Could not uniquely identify statistic column in p.table.")

param_tbl <- ptab |>
rename(
estimate  = Estimate,
std_error = `Std. Error`,
statistic = all_of(stat_col),
p_value   = all_of(p_col)
) |>
mutate(component = "parametric", model = model_name) |>
select(model, component, term, estimate, std_error, statistic, p_value) |>
mutate(across(where(is.numeric), ~ round(.x, digits)))

stab <- as.data.frame(sm$s.table)
stab <- rownames_to_column(stab, "term")
stab <- as_tibble(stab)

refdf_col <- names(stab)[grepl("^Ref\\.df$", names(stab))]
edf_col   <- names(stab)[grepl("^edf$", names(stab))]
pval_col  <- names(stab)[grepl("^p\\-value$", names(stab))]
stat_col2 <- names(stab)[grepl("^(F|Chi\\.sq)$", names(stab))]

if (length(edf_col) != 1) stop("Could not uniquely identify edf column in s.table.")
if (length(refdf_col) != 1) stop("Could not uniquely identify Ref.df column in s.table.")
if (length(pval_col) != 1) stop("Could not uniquely identify p-value column in s.table.")
if (length(stat_col2) != 1) stop("Could not uniquely identify test statistic column in s.table.")

smooth_tbl <- stab |>
rename(
edf = all_of(edf_col),
ref_df = all_of(refdf_col),
statistic = all_of(stat_col2),
p_value = all_of(pval_col)
) |>
mutate(component = "smooth", model = model_name) |>
select(model, component, term, edf, ref_df, statistic, p_value) |>
mutate(across(where(is.numeric), ~ round(.x, digits)))

combined_tbl <- bind_rows(
param_tbl |>
mutate(edf = NA_real_, ref_df = NA_real_) |>
select(model, component, term, estimate, std_error, edf, ref_df, statistic, p_value),
smooth_tbl |>
mutate(estimate = NA_real_, std_error = NA_real_) |>
select(model, component, term, estimate, std_error, edf, ref_df, statistic, p_value)
)

list(parametric = param_tbl, smooth = smooth_tbl, combined = combined_tbl)
}

occ_tabs  <- gam_results_table(final_occ,      "Occupancy", digits = 3)
cpue_tabs <- gam_results_table(final_cpue_pos, "CPUE (pos)", digits = 3)
bpue_tabs <- gam_results_table(final_BPUE_pos, "BPUE (pos)", digits = 3)

add_sig <- function(df) {
df |>
mutate(
sig = case_when(
is.na(p_value) ~ "",
p_value < 0.001 ~ "***",
p_value < 0.01  ~ "**",
p_value < 0.05  ~ "*",
p_value < 0.1   ~ ".",
TRUE ~ ""
)
)
}

occ_table  <- add_sig(occ_tabs$combined)
cpue_table <- add_sig(cpue_tabs$combined)
bpue_table <- add_sig(bpue_tabs$combined)

write.csv(occ_table,  file = file.path(out_dir, "table_gam_occupancy.csv"), row.names = FALSE)
write.csv(cpue_table, file = file.path(out_dir, "table_gam_cpue_pos.csv"),  row.names = FALSE)
write.csv(bpue_table, file = file.path(out_dir, "table_gam_bpue_pos.csv"),  row.names = FALSE)

all_models_table <- bind_rows(occ_table, cpue_table, bpue_table)
write.csv(all_models_table, file = file.path(out_dir, "table_gam_all_models.csv"), row.names = FALSE)

all_models_table |>
  mutate(
    term = gsub("s\\(|\\)", "", term),
    p_value = ifelse(!is.na(p_value), format.pval(p_value, digits = 3, eps = 0.001), NA)
  ) |>
  select(-model) |>
  kable(
    caption = "GAM model results for kōura occupancy, CPUE, and BPUE models. Significance codes: *** p < 0.001, ** p < 0.01, * p < 0.05, . p < 0.1",
    col.names = c("Component", "Term", "Estimate", "SE", "EDF", "Ref.df", "Statistic", "p-value", ""),
    align = c("l","l","r","r","r","r","r","r","l")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  kableExtra::group_rows("Occupancy", 1, nrow(occ_table)) |>
  kableExtra::group_rows("CPUE (positive sites)", nrow(occ_table) + 1, nrow(occ_table) + nrow(cpue_table)) |>
  kableExtra::group_rows("BPUE (positive sites)", nrow(occ_table) + nrow(cpue_table) + 1, nrow(all_models_table))
```

# PCA (habitat complexity)
```{r}
# Comment Frank about the modeling:
  # If you centered (on the mean) and standardized (scaled to unit variance) the predictors you could use the parameter estimates as an effect size to say something about the importance of the predictors (i.e. weak vs strong effects).

#| label: pca

habitat_vars <- M_C_data %>%
dplyr::select(
Riparian_vegetation,
Substrate_index,
Wood_cover,
Overhanging_trees,
Emergent_Native,
Submerged_Native,
Submerged_Non_Native
) %>%
dplyr::filter(complete.cases(.))

habitat_pca <- prcomp(habitat_vars, center = TRUE, scale. = TRUE)

summary(habitat_pca)
biplot(habitat_pca)
habitat_pca$rotation


```

# Koura catfish interaction
```{r}
# 1. Catfish distribution across habitat types
M_C_data %>%
  group_by(Habitat_Type) %>%
  summarise(n_sites = n(),
            catfish_present = sum(Presence_Catfish, na.rm = TRUE),
            koura_present = sum(Presence_Kōura, na.rm = TRUE),
            mean_CPUE_koura = mean(Weighted_CPUE_Kōura, na.rm = TRUE))

# 2. Kōura metrics: catfish-present vs absent sites
M_C_data %>%
  group_by(Presence_Catfish) %>%
  summarise(n = n(),
            koura_presence_rate = mean(Presence_Kōura, na.rm = TRUE),
            mean_CPUE_koura = mean(Weighted_CPUE_Kōura, na.rm = TRUE),
            mean_BPUE_koura = mean(Weighted_BPUE_Kōura, na.rm = TRUE))

# 3. Kōura metrics: catfish-invaded vs catfish-free lakes
M_C_data %>%
  mutate(catfish_lake = if_else(Lake %in% c("Rotorua", "Rotoiti"),
                                "Catfish present", "Catfish absent")) %>%
  group_by(catfish_lake) %>%
  summarise(n_sites = n(),
            koura_presence_rate = mean(Presence_Kōura, na.rm = TRUE),
            mean_CPUE_koura = mean(Weighted_CPUE_Kōura, na.rm = TRUE),
            mean_BPUE_koura = mean(Weighted_BPUE_Kōura, na.rm = TRUE))

```

