---
title: "Complex littoral habitats enhance freshwater crayfish populations"
author: "Olivier V. Raven"
format:
  html:
    toc: true
    number-sections: true
---

## Overview
This site accompanies the scientific work on: 'Complex littoral habitats enhance freshwater crayfish populations in five volcanic lakes
of Aotearoa New Zealand'

- **Code repository:** (add URL)
- **Data DOI:** (add DOI link)
- **Paper DOI:** (add DOI link)
- **title:** Complex littoral habitats enhance freshwater crayfish populations in five volcanic lakes
of Aotearoa New Zealand

# Load packages
```{r}
cat("\014"); rm(list = ls())

# Define the list of packages
packages <- c("emmeans","rstatix", "multcompView","brglm2","patchwork", "gratia","pROC","caret", "mgcv","tidyr","purrr", "corrplot","glmmTMB", "performance","tibble", "tidyverse", "dplyr","writexl", "readxl")

# Load packages if not already installed
quiet_load <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    suppressWarnings(suppressMessages(install.packages(pkg, dependencies = TRUE)))}
  suppressPackageStartupMessages(require(pkg, character.only = TRUE, quietly = TRUE))
  invisible(TRUE)}

invisible(lapply(packages, quiet_load))

```

# Load data
```{r}
getwd()
exc_file_dir <- "data/raw/Natural_habitat.xlsx"
raw_data_dir <- "data/raw"
der_data_dir <- "data/derived"
fig_dir      <- "figures"


# Import the data sets 
Site_info          <- read_excel(exc_file_dir, sheet = "Site_info")
Monitoring_data    <- read_excel(exc_file_dir, sheet = "Monitoring_data")
Weed_data          <- read_excel(exc_file_dir, sheet = "Weed_data")  %>% dplyr::select(-starts_with("..."))
Fish_data          <- read_excel(exc_file_dir, sheet = "Fish_data") %>% dplyr::select(-starts_with("..."))
Macroinvertebrates <- read_excel(exc_file_dir, sheet = "Macroinvertebrates")

# filter for only using each site 1 time
Site_info          <- Site_info |> dplyr::filter(!stringr::str_ends(Monitoring_ID, "_1"))
Monitoring_data    <- Monitoring_data |> dplyr::filter(!stringr::str_ends(Monitoring_ID, "_1"))
Weed_data          <- Weed_data |> dplyr::filter(!stringr::str_ends(Monitoring_ID, "_1"))
Fish_data          <- Fish_data |> dplyr::filter(!stringr::str_ends(Monitoring_ID, "_1"))
Macroinvertebrates <- Macroinvertebrates |> dplyr::filter(!stringr::str_ends(Monitoring_ID, "_1"))

```

# Calculte CPUE & BPUE
```{r fig.width=8, fig.height=5, dpi=100, echo=FALSE, message=FALSE, warning=FALSE}
# use size of koura to fill missing weights.
lw_model <- lm(log10(Weight_g) ~ log10(Length_mm),
               data = Fish_data,
               subset = Species_name == "Freshwater_crayfish" | Maori_name == "Kōura",
               na.action = na.exclude)

# extract the residual variance
sigma_log10 <- sigma(lw_model)
c <- 10^(0.5 * sigma_log10^2)

Fish_data <- Fish_data |>
  mutate(is_koura = Species_name == "Freshwater_crayfish" | Maori_name == "Kōura",
    Predicted_weight = case_when( !is.na(Weight_g) ~ Weight_g,
      is_koura ~ {pred_log <- predict(lw_model, newdata = pick(everything()))
        10^(pred_log) * c}, TRUE ~ NA_real_ ),
    Weight_source = case_when(is_koura & !is.na(Weight_g) ~ "measured",
      is_koura & is.na(Weight_g) ~ "predicted",TRUE ~ NA_character_))

count_df <- Fish_data |>
  dplyr::filter(is_koura, !is.na(Weight_source)) |>
  dplyr::count(Weight_source, name = "n")

measured_n  <- count_df$n[count_df$Weight_source == "measured"]
predicted_n <- count_df$n[count_df$Weight_source == "predicted"]

total_koura <- Fish_data |>
  dplyr::filter(is_koura) |>
  nrow()

hist(Fish_data |> dplyr::filter(is_koura) |> dplyr::pull(Length_mm),xlab = "Length (mm)")

# Extract model coefficients
a <- coef(lw_model)[1]
b <- coef(lw_model)[2]

# Create formula text for the plot
formula_text <- paste0( "log10(Weight[g]) = ", round(a, 3), " + ", round(b, 3), " * log10(OCL Length[mm])")

# Length–weight relationship plot (Springer-compliant)
lengt_weight_plot <- ggplot(
  Fish_data |> dplyr::filter(is_koura),
  aes(x = Length_mm, y = Predicted_weight, shape = Weight_source)) +
  geom_point(size = 2.2, colour = "black", stroke = 0.6) +
  scale_shape_manual(values = c("measured" = 16, "predicted" = 1)) +
  labs(x = "OCL length (mm)", y = "Weight (g)", shape = "Weight source") +
  annotate("text", x = Inf, y = Inf, label = formula_text, hjust = 1.1, vjust = 1.3, size = 3) +
  theme_classic() +
  theme(text = element_text(family = "Arial", size = 8),
    axis.title = element_text(face = "plain"),
    axis.text = element_text(face = "plain"),
    legend.title = element_text(face = "plain"),
    legend.text = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3))

lengt_weight_plot

ggsave(filename = file.path(fig_dir, "lengt_weight_plot.png"), plot = lengt_weight_plot, width = 170, height = 120, units = "mm", dpi = 1200)


CPUE_BPUE_legacy <- Fish_data %>%
  filter(!is.na(Species)) %>%
  group_by(Monitoring_ID, Species, Net_type) %>%
  dplyr::reframe(
    Total_Individuals = sum(Amount, na.rm = TRUE),
    Total_Weight      = sum(Predicted_weight, na.rm = TRUE),
    Total_Effort      = dplyr::first(Amount_nets),
    CPUE              = Total_Individuals / Total_Effort,
    BPUE              = Total_Weight      / Total_Effort,
    Mean_Length       = mean(Length_mm, na.rm = TRUE),
    Min_Length        = ifelse(all(is.na(Length_mm)), NA, min(Length_mm, na.rm = TRUE)),
    Max_Length        = ifelse(all(is.na(Length_mm)), NA, max(Length_mm, na.rm = TRUE)),
    Mean_Weight       = mean(Predicted_weight, na.rm = TRUE),
    Min_Weight        = ifelse(all(is.na(Predicted_weight)), NA, min(Predicted_weight, na.rm = TRUE)),
    Max_Weight        = ifelse(all(is.na(Predicted_weight)), NA, max(Predicted_weight, na.rm = TRUE)))

CPUE_BPUE_weighted <- CPUE_BPUE_legacy %>%
  group_by(Monitoring_ID, Species) %>%
  summarise(
    Total_Individuals       = sum(Total_Individuals, na.rm = TRUE),
    Total_Weight            = sum(Total_Weight,      na.rm = TRUE),
    Weighted_CPUE_numerator = sum(CPUE * Total_Effort,  na.rm = TRUE),
    Weighted_BPUE_numerator = sum(BPUE * Total_Effort,  na.rm = TRUE),
    Total_Effort_sum        = sum(Total_Effort,         na.rm = TRUE),
    Mean_Length             = mean(Mean_Length, na.rm = TRUE),
    Min_Length              = ifelse(all(is.na(Min_Length)), NA, min(Min_Length, na.rm = TRUE)),
    Max_Length              = ifelse(all(is.na(Max_Length)), NA, max(Max_Length, na.rm = TRUE)),
    Mean_Weight             = mean(Mean_Weight, na.rm = TRUE),
    Min_Weight              = ifelse(all(is.na(Min_Weight)), NA, min(Min_Weight, na.rm = TRUE)),
    Max_Weight              = ifelse(all(is.na(Max_Weight)), NA, max(Max_Weight, na.rm = TRUE)),
    .groups = "drop") %>%
  ungroup() %>%
  mutate(Total_Effort_sum = ifelse(Monitoring_ID %in% c("96_0", "101_0", "117_1", "119_1"), 3, 4),
    Weighted_CPUE    = Weighted_CPUE_numerator / Total_Effort_sum,
    Weighted_BPUE    = Weighted_BPUE_numerator / Total_Effort_sum)

species_presence_absence <- Fish_data %>%
  filter(!is.na(Species)) %>%
  distinct(Monitoring_ID, Species) %>%
  mutate(Presence = 1) %>%
  pivot_wider(names_from  = Species, values_from = Presence,  values_fill = list(Presence = 0), names_prefix = "Presence_") %>%
  mutate(Predator_Fish_Presence = pmax(Presence_Trout, Presence_Eel, Presence_Catfish))

CPUE_BPUE_weighted_summary <- CPUE_BPUE_weighted %>%
  pivot_wider(names_from  = Species,
    values_from = c(Total_Individuals, Weighted_CPUE, Weighted_BPUE, Total_Weight,
      Mean_Length, Mean_Weight, Weighted_CPUE_numerator, Weighted_BPUE_numerator,
      Total_Effort_sum, Min_Length, Max_Length, Min_Weight, Max_Weight), names_sep   = "_",
    values_fill = list(Total_Individuals = 0, Weighted_CPUE = 0, Weighted_BPUE = 0)) %>%
  mutate(Richness  = rowSums(dplyr::select(., starts_with("Total_Individuals_")) > 0),
    Abundance = rowSums(dplyr::select(., starts_with("Total_Individuals_") & !ends_with(c("_Bullies", "_Common_smelt")))))

```

# Combine the data frames
```{r}
unit_metadata <- Monitoring_data %>%
  dplyr::select(Parameter, Unit) %>%
  distinct()

Monitoring_summary <- Monitoring_data %>%
  dplyr::select(-Group, -Notes, -Unit) %>%
  pivot_wider(
    names_from  = c(Parameter),
    values_from = Value,
    values_fill = list(Value = NA))

Monitoring_summary <- Monitoring_summary %>%
  mutate(across(c(
      Bottom_visible, Water_clarity, Depth_10m, Slope, Riparian_vegetation, Vegetation_nearby,
      Overhanging_trees, Erosion, Sructure, Bedrock, Boulders, Cobble, Gravel,
      Sand, Mud, Organic_matter, Rock_size, Temperature, DO_mgl, DO_percent,
      Conductivity, Specific_conductivity, pH, Wood_cover), ~ as.numeric(.)))

Weed_summary <- Weed_data %>%
  group_by(Monitoring_ID, Weed_Type, Native_Status) %>%
  summarise(Total_Cover = sum(Percentage_Cover, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from  = c(Weed_Type, Native_Status), values_from = Total_Cover,values_fill = 0)

Macroinvertebrates_sum <- Macroinvertebrates %>%
  group_by(Monitoring_ID, Species) %>%
  summarise(Total_amount = sum(Amount, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = c(Species), values_from = Total_amount, values_fill = 0)

Macroinvertebrates_sum <- Macroinvertebrates %>%
  group_by(Monitoring_ID, Species) %>%
  summarise(Total_amount = sum(Amount, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = c(Species), values_from = Total_amount, values_fill = 0) %>%
  mutate(
    Invertebrates_Richness  = rowSums(dplyr::select(., -Monitoring_ID) > 0),
    Invertebrates_Abundance = rowSums(dplyr::select(., -Monitoring_ID)))

Monitoring_CPUE_data <- Site_info %>%
  left_join(Monitoring_summary, by = "Monitoring_ID") %>%
  left_join(Weed_summary %>% dplyr::select(Monitoring_ID,
                                           Emergent_Native, Emergent_Non_Native,
                                           Submerged_Native, Submerged_Non_Native, Turf_Native),
            by = c("Monitoring_ID")) %>%
  left_join(CPUE_BPUE_weighted_summary, by = "Monitoring_ID") %>%
  left_join(species_presence_absence, by = "Monitoring_ID") %>%
  left_join(Macroinvertebrates_sum, by = "Monitoring_ID")

Monitoring_CPUE_data <- Monitoring_CPUE_data %>%
  mutate(
    Presence_rocks = if_else(Cobble > 1 | Boulders > 1, 1, 0),
    Slope_5m       = 5 / Distance_5m,
    Site_ID_       = Site_ID.x - 60,
    Monitoring_ID_ = paste0(as.numeric(sub("_.*", "", Monitoring_ID)) - 60,
                            sub("^[^_]*", "", Monitoring_ID)),
    Monitoring     = sub(".*?_", "", Monitoring_ID),
    Date           = as.Date(Date_Time),
    Time           = format(Date_Time, "%H:%M:%S"),
    Year           = lubridate::year(Date_Time),
    Month          = lubridate::month(Date_Time, label = TRUE),
    Day            = lubridate::day(Date_Time),
    Season         = case_when(
      Month %in% c("Dec", "Jan", "Feb") ~ "Summer",
      Month %in% c("Mar", "Apr", "May") ~ "Autumn",
      Month %in% c("Jun", "Jul", "Aug") ~ "Winter",
      Month %in% c("Sep", "Oct", "Nov") ~ "Spring",
      TRUE ~ NA_character_),Date_Time_Numeric = as.numeric(Date_Time))

habitat_classification <- Monitoring_CPUE_data %>%
  dplyr::select(
    Monitoring_ID, DHT, Lake,
    Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter,
    Emergent_Native, Emergent_Non_Native,
    Submerged_Native, Submerged_Non_Native, Wood_cover) %>%
  pivot_longer(
    cols = c(Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter,
      Emergent_Native, Emergent_Non_Native,
      Submerged_Native, Submerged_Non_Native, Wood_cover    ),
    names_to = "Type",values_to = "Percentage") %>%
  group_by(Monitoring_ID) %>%
  summarise(Rocky_Percentage = sum(Percentage[Type %in% c("Bedrock", "Boulders", "Cobble")],    na.rm = TRUE),
    Sand_Percentage = sum(Percentage[Type == "Sand"],  na.rm = TRUE),
    Mud_Percentage = sum(Percentage[Type %in% c("Mud", "Organic_matter")],na.rm = TRUE),
    Emergent_Percentage = sum(Percentage[Type == "Emergent_Native"],na.rm = TRUE),
    Substrate_index = sum(
      0.08 * Percentage[Type == "Bedrock"] +
      0.07 * Percentage[Type == "Boulders"] +
      0.06 * Percentage[Type == "Cobble"] +
      0.04 * Percentage[Type == "Gravel"] +
      0.03 * Percentage[Type == "Sand"] +
      0.02 * Percentage[Type == "Organic_matter"] +
      0.01 * Percentage[Type == "Mud"],
      na.rm = TRUE),
    Substrate_CV = {
      substrate_vals <- Percentage[
        Type %in% c("Bedrock", "Boulders", "Cobble", "Gravel",
                    "Sand", "Mud", "Organic_matter")]
      substrate_vals <- substrate_vals[!is.na(substrate_vals)]
      if (length(substrate_vals) > 1 && mean(substrate_vals) > 0) {
        sd(substrate_vals) / mean(substrate_vals)
      } else {
        NA_real_
      }
    },
    .groups = "drop"  ) %>%
  mutate(
    Habitat_Type = case_when(
      Rocky_Percentage > 25 ~ "Rocky",
      Emergent_Percentage > 25 ~ "Emergent Macrophyte",
      Sand_Percentage >= Mud_Percentage ~ "Sandy",
      TRUE ~ "Muddy")) %>%
  dplyr::select(
    Monitoring_ID,
    Habitat_Type,
    Substrate_index,
    Substrate_CV)

cor(habitat_classification$Substrate_index,
    habitat_classification$Substrate_CV,
    use = "complete.obs")


Monitoring_CPUE_data <- Monitoring_CPUE_data %>%
  left_join(habitat_classification, by = c("Monitoring_ID"))

writexl::write_xlsx(Monitoring_CPUE_data, file.path(der_data_dir, "Monitoring_CPUE_data.xlsx"))
write.csv(Monitoring_CPUE_data, file.path(der_data_dir, "Monitoring_CPUE_data.csv"), row.names = FALSE)
write.csv(habitat_classification, file.path(der_data_dir, "habitat_classification.csv"), row.names = FALSE)

list(habitat_classification_head = head(habitat_classification, 5),monitoring_cpue_head        = head(Monitoring_CPUE_data, 5))

```

# Plot all variables 
```{r fig.width=10, fig.height=10, dpi=100}
# set orders
M_C_data <- Monitoring_CPUE_data
lake_order <- c("Rotorua", "Rotoiti", "Rotoehu", "Rotomā", "Ōkāreka")
M_C_data$Lake <- factor(M_C_data$Lake, levels = lake_order)

# nice facet names
var_lab <- c(
  Substrate_index       = "Substrate index",
  Slope_5m              = "Slope to 5 m depth",
  Riparian_vegetation   = "Riparian vegetation (%)",
  Overhanging_trees     = "Overhanging trees (%)",
  Wood_cover            = "Wood cover (%)",
  Emergent_vegetation   = "Emergent vegetation (%)",
  Submerged_vegetation  = "Submerged vegetation (%)",
  Temperature           = "Surface temperature (°C)",
  DO_mgl                = "Dissolved oxygen (mg l-1)",
  Specific_conductivity = "Specific conductivity (uS cm-1)",
  pH                    = "pH",
  Presence_Catfish       = "Brown bullhead catfish presence (proportion of sites)",
  Presence_Eel           = "Eel presence (proportion of sites)",
  Presence_Goldfish      = "Goldfish presence (proportion of sites)",
  Presence_Common_smelt  = "Common smelt presence (proportion of sites)",
  Presence_Kōura         = "Kōura presence (proportion of sites)",
  Presence_Kōaro         = "Kōaro presence (proportion of sites)",
  Presence_Bullies       = "Bullies presence (proportion of sites)",
  Presence_Trout         = "Trout presence (proportion of sites)",
  Presence_Mosquitofish  = "Mosquitofish presence (proportion of sites)")
  
 

# env (boxplots)
Env_data <- M_C_data %>%
  mutate(
    Emergent_vegetation  = Emergent_Native + Emergent_Non_Native,
    Submerged_vegetation = Submerged_Native + Submerged_Non_Native) %>%
  select(
    Lake, Substrate_index, Slope_5m, Riparian_vegetation, Overhanging_trees, Wood_cover,
    Temperature, DO_mgl, pH, Specific_conductivity, Emergent_vegetation, Submerged_vegetation) %>%
  pivot_longer(-Lake, names_to = "Variable", values_to = "Value") %>%
  mutate(Var_type = "Env")

# fish (bars)
Fish_data <- M_C_data %>%
  select(Lake, Presence_Eel, Presence_Common_smelt, Presence_Catfish, Presence_Goldfish) %>%
  pivot_longer(-Lake, names_to = "Variable", values_to = "Presence") %>%
  group_by(Lake, Variable) %>%
  summarise(Value = mean(Presence, na.rm = TRUE), .groups = "drop") %>%
  mutate(Var_type = "Fish")

EnvBio_data <- bind_rows(
  Env_data %>% select(Lake, Variable, Value, Var_type),
  Fish_data) %>%
  mutate(
    Lake = factor(Lake, levels = lake_order),
    Variable = factor(Variable, levels = names(var_lab)))


# make into table 
# helper for 95% CI (t-based)
ci95 <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 2) return(c(NA_real_, NA_real_))
  se <- sd(x) / sqrt(n)
  tcrit <- qt(0.975, df = n - 1)
  m <- mean(x)
  c(m - tcrit * se, m + tcrit * se)
}

# 4a. Env variables summary per lake (site-level)
Env_summary_table <- Env_data %>%
  group_by(Lake, Variable) %>%
  summarise(
    n       = sum(!is.na(Value)),
    Mean    = mean(Value, na.rm = TRUE),
    Median  = median(Value, na.rm = TRUE),
    Min     = min(Value, na.rm = TRUE),
    Max     = max(Value, na.rm = TRUE),
    CI_low  = ci95(Value)[1],
    CI_high = ci95(Value)[2],
    .groups = "drop") %>%
  mutate(Var_type = "Env")

# Env: add "All lakes"
Env_summary_all <- Env_data %>%
  group_by(Variable) %>%
  summarise(
    n       = sum(!is.na(Value)),
    Mean    = mean(Value, na.rm = TRUE),
    Median  = median(Value, na.rm = TRUE),
    Min     = min(Value, na.rm = TRUE),
    Max     = max(Value, na.rm = TRUE),
    CI_low  = ci95(Value)[1],
    CI_high = ci95(Value)[2],
    .groups = "drop"  ) %>%
  mutate(Lake = "All lakes", Var_type = "Env"  )

# 4b. Fish presence summary per lake (site-level)
Fish_summary_table <- M_C_data %>%
  select(Lake, Presence_Eel, Presence_Common_smelt, Presence_Catfish,
         Presence_Goldfish, Presence_Kōaro, Presence_Trout) %>%
  pivot_longer(-Lake, names_to = "Variable", values_to = "Presence") %>%
  group_by(Lake, Variable) %>%
  summarise(
    n       = sum(!is.na(Presence)),
    Mean    = mean(Presence, na.rm = TRUE),
    Median  = median(Presence, na.rm = TRUE),
    Min     = min(Presence, na.rm = TRUE),
    Max     = max(Presence, na.rm = TRUE),
    CI_low  = ci95(Presence)[1],
    CI_high = ci95(Presence)[2],
    .groups = "drop") %>%
  mutate(Var_type = "Fish")

# Fish: add "All lakes" (report only overall proportion + CI; other stats not informative)
Fish_summary_all <- M_C_data %>%
  select(Presence_Eel, Presence_Common_smelt, Presence_Catfish,
         Presence_Goldfish, Presence_Kōaro, Presence_Trout) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Presence") %>%
  group_by(Variable) %>%
  summarise(
    n       = sum(!is.na(Presence)),
    Mean    = mean(Presence, na.rm = TRUE),
    CI_low  = ci95(Presence)[1],
    CI_high = ci95(Presence)[2],
    .groups = "drop"  ) %>%
  mutate(Lake = "All lakes",
    Var_type = "Fish",
    Median = NA_real_,
    Min    = NA_real_,
    Max    = NA_real_  ) %>%
  select(Lake, Variable, Var_type, n, Mean, Median, Min, Max, CI_low, CI_high)

# 4c. Combined summary table (Env + Fish, incl. All lakes)
EnvBio_summary_table <- bind_rows(
  Env_summary_table,
  Env_summary_all,
  Fish_summary_table,
  Fish_summary_all) %>%
  select(Lake, Variable, Var_type, n, Mean, Median, Min, Max, CI_low, CI_high) %>%
  mutate(Variable = factor(Variable, levels = levels(EnvBio_data$Variable)),
    Lake = factor(Lake, levels = c(lake_order, "All lakes"))  ) %>%
  arrange(Variable, Lake)

EnvBio_summary_table

#write.csv(EnvBio_summary_table, file = file.path(fig_dir, "EnvBio_summary_table.csv"),  row.names = FALSE)
```

# Kōura presence, CPUE, and BPUE plot 
```{r fig.width=6, fig.height=6, dpi=100}
plot_koura_stats <- function(data, y_var, y_label,
                             type = c("continuous", "presence"),
                             show_x_title = FALSE) {

  type <- match.arg(type)
  lakes <- levels(data$Lake)
  xlab_text <- if (show_x_title) "Lake" else NULL

  base_theme <- theme_classic() +
    theme(
      text = element_text(family = "Arial", size = 8),
      axis.title = element_text(face = "plain"),
      axis.text = element_text(face = "plain"),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3)
    )

  if (type == "continuous") {

    ggplot(data, aes(Lake, .data[[y_var]])) +
      geom_boxplot(
        fill = NA,
        colour = "black",
        linewidth = 0.3,
        outlier.size = 1.5
      ) +
      labs(y = y_label, x = xlab_text) +
      base_theme

  } else {

    presence_summary <- data %>%
      group_by(Lake) %>%
      summarise(
        n = sum(!is.na(.data[[y_var]])),
        k = sum(.data[[y_var]] == 1, na.rm = TRUE),
        Presence_Rate = ifelse(n > 0, k / n, NA_real_),
        .groups = "drop"
      ) %>%
      rowwise() %>%
      mutate(
        CI_low  = ifelse(n > 0, binom.test(k, n)$conf.int[1], NA_real_),
        CI_high = ifelse(n > 0, binom.test(k, n)$conf.int[2], NA_real_)
      ) %>%
      ungroup() %>%
      mutate(Lake = factor(Lake, levels = lakes))

    ggplot(presence_summary, aes(x = Lake, y = Presence_Rate)) +
      geom_col(
        fill = NA,
        colour = "black",
        linewidth = 0.3
      ) +
      geom_errorbar(
        aes(ymin = CI_low, ymax = CI_high),
        width = 0.15,
        linewidth = 0.3
      ) +
      scale_y_continuous(limits = c(0, 1)) +
      labs(x = xlab_text, y = y_label) +
      base_theme
  }
}

KPRES_plot <- plot_koura_stats(M_C_data, "Presence_Kōura", "Kōura presence", type = "presence", show_x_title = FALSE)
KCPUE_plot <- plot_koura_stats(M_C_data, "Weighted_CPUE_Kōura", "Kōura CPUE", type = "continuous", show_x_title = FALSE)
KBPUE_plot <- plot_koura_stats(M_C_data, "Weighted_BPUE_Kōura", "Kōura BPUE", type = "continuous", show_x_title = TRUE)

Koura_plots <- KPRES_plot / KCPUE_plot / KBPUE_plot
Koura_plots

ggsave(file.path(fig_dir, "Koura_plots.png"), Koura_plots, width = 120, height = 120, units = "mm", dpi = 1200)

# model fitting 
fit_presence_glmm <- function(data, response, lake_var = "Lake", random_effect = "Habitat_Type") {
  # Binomial GLMM: presence ~ Lake + (1|Habitat_Type)
  # If you have repeated sampling of sites, use a site ID random effect as well.
  form <- as.formula(paste0(response, " ~ ", lake_var, " + (1|", random_effect, ")"))
  glmmTMB(form, data = data, family = binomial(link = "logit"))
}

fit_tweedie_glmm <- function(data, response, lake_var = "Lake", random_effect = "Habitat_Type") {
  # Tweedie GLMM: response ~ Lake + (1|Habitat_Type)
  # Tweedie is suitable for non-negative continuous with many zeros / right skew.
  form <- as.formula(paste0(response, " ~ ", lake_var, " + (1|", random_effect, ")"))
  glmmTMB(form, data = data, family = tweedie(link = "log"))
}

# Presence (binomial GLMM)
m_koura_pres <- fit_presence_glmm(M_C_data, response = "Presence_Kōura", lake_var = "Lake", random_effect = "Habitat_Type")

# CPUE (Tweedie GLMM)
m_koura_cpue <- fit_tweedie_glmm(M_C_data, response = "Weighted_CPUE_Kōura", lake_var = "Lake", random_effect = "Habitat_Type")

# BPUE (Tweedie GLMM)
m_koura_bpue <- fit_tweedie_glmm(M_C_data, response = "Weighted_BPUE_Kōura", lake_var = "Lake", random_effect = "Habitat_Type")

# ---- model outputs ----

# Overall lake effects (Type II/III tests are not native in glmmTMB; use drop1 LRT)
drop1(m_koura_pres, test = "Chisq")
drop1(m_koura_cpue, test = "Chisq")
drop1(m_koura_bpue, test = "Chisq")

# Pairwise lake comparisons (model-based; BH-adjusted)
pres_pairs <- emmeans(m_koura_pres, ~ Lake, type = "response") %>%
  pairs(adjust = "BH")

cpue_pairs <- emmeans(m_koura_cpue, ~ Lake, type = "response") %>%
  pairs(adjust = "BH")

bpue_pairs <- emmeans(m_koura_bpue, ~ Lake, type = "response") %>%
  pairs(adjust = "BH")

pres_pairs
cpue_pairs
bpue_pairs
```

# GAM helpers 
```{r}
# GAM Modeling setup & helpers
Modeling_data <- Monitoring_CPUE_data 

alpha_sig    <- 0.1      # significance threshold for showing fish plots
p_cutoff_ml  <- 0.05      # stepwise ML drop threshold
vif_thresh   <- 5
INCLUDE_RE   <- TRUE      # use LID as random effect

custom_k <- list(
  Slope_5m           = 10,
  Riparian_vegetation= 7,
  Overhanging_trees  = 5,
  Wood_cover         = 10,
  Substrate_index    = 10,
  Temperature        = 10,
  pH                 = 10,
  DO_mgl             = 10,
  Emergent_Native    = 9,
  Submerged_Non_Native = 9,
  Turf_Native        = 6,
  Submerged_Native   = 6
)

fish_vars <- c(
  "Presence_Goldfish",
  "Presence_Eel",
  "Presence_Catfish",
  "Presence_Common_smelt",
  "Presence_Kōaro"
)

vars_common <- c(
  "LID",
  "Slope_5m","Riparian_vegetation","Overhanging_trees","Wood_cover","Substrate_index",
  "Temperature","pH","DO_mgl","DO_percent","Specific_conductivity",
  "Emergent_Native","Submerged_Non_Native","Turf_Native","Submerged_Native","Emergent_Non_Native",
  fish_vars
)

# ---------- Generic helpers ----------

is_cont <- function(x) is.numeric(x) && dplyr::n_distinct(x, na.rm = TRUE) >= 5

# Construct a smooth or parametric term string for a single variable
build_smooth_gam <- function(var, data, custom_k = list(), include_re_for_LID = FALSE) {
  # LID as random effect if requested
  if (identical(var, "LID") && include_re_for_LID) return("s(LID, bs='re')")
  x <- data[[var]]
  if (is.numeric(x)) {
    nuniq <- dplyr::n_distinct(x, na.rm = TRUE)
    if (nuniq < 5) return(var)  # treat binary/few-level numerics as parametric
    k_req <- if (!is.null(custom_k[[var]])) custom_k[[var]] else 10
    k_cap <- max(3, min(k_req, nuniq - 1))
    return(paste0("s(", var, ", bs='ts', k=", k_cap, ")"))
  }
  var
}

# Helper to exclude random effect smooths (currently only LID) from prediction
exclude_if_RE <- function(mod){
  if (!length(mod$smooth)) return(NULL)
  has_re <- vapply(mod$smooth, function(s) "LID" %in% s$term, logical(1))
  if (any(has_re)) "s(LID)" else NULL
}

# ---------- VIF helpers ----------

# VIF pruning (binomial) using brglm2 to reduce separation issues
remove_high_vif_glmBI <- function(
  data,
  response,
  predictors,
  threshold    = vif_thresh,
  protect_vars = character(0),
  verbose      = TRUE
) {
  # 1) Remove near-zero variance predictors
  nzv <- caret::nearZeroVar(data[, predictors, drop = FALSE])
  if (length(nzv)) {
    if (verbose) {
      message("Removing near-zero variance predictors: ",
              paste(predictors[nzv], collapse = ", "))
    }
    predictors <- predictors[-nzv]
  }

  rpt <- list(
    removed = character(),
    start   = predictors
  )

  last_vif <- NULL

  repeat {
    if (!length(predictors)) {
      stop("All predictors removed during VIF pruning.")
    }

    # Build formula for current set of predictors
    fml <- as.formula(paste(response, "~", paste(predictors, collapse = " + ")))

    # Fit standard lm 
    model <- try(
  lm(fml, data = data),
  silent = TRUE
)

    if (inherits(model, "try-error")) {
      if (verbose) {
        message("Model failed to fit during VIF pruning. Stopping.\n",
                "Error was: ", attr(model, "condition")$message)
      }
      break
    }

    # Compute VIF
    vif_data <- performance::check_collinearity(model)
    vif_data <- vif_data[!grepl("\\|", vif_data$Term), , drop = FALSE]

    last_vif <- vif_data

    # If no rows or all VIFs below threshold, we're done
    if (!nrow(vif_data) || all(vif_data$VIF < threshold)) {
      if (verbose) {
        message("All remaining predictors have VIF < ", threshold, ".")
      }
      break
    }

    # Order by VIF descending and pick the first non-protected variable
    ord <- order(vif_data$VIF, decreasing = TRUE)
    to_remove <- NA_character_

    for (i in ord) {
      cand <- vif_data$Term[i]
      if (!(cand %in% protect_vars)) {
        to_remove <- cand
        break
      }
    }

    # If all high-VIF variables are protected, stop
    if (is.na(to_remove)) {
      if (verbose) {
        message("All high-VIF variables are in protect_vars; stopping pruning.")
      }
      break
    }

    if (verbose) {
      this_vif <- vif_data$VIF[vif_data$Term == to_remove]
      message("Removing '", to_remove,
              "' with VIF = ", signif(this_vif, 4))
    }

    predictors   <- setdiff(predictors, to_remove)
    rpt$removed  <- c(rpt$removed, to_remove)
  }

  # Store the final VIF table in the report
  rpt$final_vif <- last_vif

  if (verbose && !is.null(last_vif)) {
    message("\nFinal VIF table for kept predictors:")
    print(last_vif)
  }

  list(
    predictors = predictors,
    report     = rpt
  )
}

# VIF pruning for non-binomial positive models (e.g. Gamma)
# Uses linear models for VIF, which is the standard and distribution-free.
remove_high_vif_glmmTMB <- function(
  data,
  response,
  predictors,
  threshold    = vif_thresh,
  protect_vars = character(0),
  verbose      = TRUE
) {
  # 1) Remove near-zero variance predictors
  nzv <- caret::nearZeroVar(data[, predictors, drop = FALSE])
  if (length(nzv)) {
    if (verbose) {
      message("Removing near-zero variance predictors: ",
              paste(predictors[nzv], collapse = ", "))
    }
    predictors <- predictors[-nzv]
  }

  rpt <- list(
    removed = character(),
    start   = predictors
  )

  last_vif <- NULL

  repeat {
    if (!length(predictors)) {
      stop("All predictors removed during VIF pruning.")
    }

    # Build formula for current set of predictors
    fml <- as.formula(paste(response, "~", paste(predictors, collapse = " + ")))

    # Fit a basic linear model purely for VIF calculation
    model <- try(
      lm(fml, data = data),
      silent = TRUE
    )

    if (inherits(model, "try-error")) {
      if (verbose) {
        err_msg <- tryCatch(attr(model, "condition")$message,
                            error = function(e) "Unknown error")
        message("lm() failed during VIF pruning. Stopping.\n",
                "lm() error was: ", err_msg)
      }
      break
    }

    # Compute VIFs
    vif_data <- performance::check_collinearity(model)
    vif_data <- vif_data[!grepl("\\|", vif_data$Term), , drop = FALSE]

    last_vif <- vif_data

    # If no rows or all VIFs below threshold, we are done
    if (!nrow(vif_data) || all(vif_data$VIF < threshold)) {
      if (verbose) {
        message("All remaining predictors have VIF < ", threshold, ".")
      }
      break
    }

    # Order by VIF descending and pick the first non-protected variable
    ord <- order(vif_data$VIF, decreasing = TRUE)
    to_remove <- NA_character_

    for (i in ord) {
      cand <- vif_data$Term[i]
      if (!(cand %in% protect_vars)) {
        to_remove <- cand
        break
      }
    }

    # If all high-VIF variables are protected, stop
    if (is.na(to_remove)) {
      if (verbose) {
        message("All high-VIF variables are in protect_vars; stopping pruning.")
      }
      break
    }

    if (verbose) {
      this_vif <- vif_data$VIF[vif_data$Term == to_remove]
      message("Removing '", to_remove,
              "' with VIF = ", signif(this_vif, 4))
    }

    predictors  <- setdiff(predictors, to_remove)
    rpt$removed <- c(rpt$removed, to_remove)
  }

  # Store final VIF table
  rpt$final_vif <- last_vif

  if (verbose && !is.null(last_vif)) {
    message("\nFinal VIF table for kept predictors (VIF from lm):")
    print(last_vif)
  }

  list(
    predictors = predictors,
    report     = rpt
  )
}



# ---------- P-value and plotting helpers ----------

# Extract p-value for either parametric var or smooth s(var)
get_term_p <- function(m, var, kind = c("any","param","smooth")){
  kind <- match.arg(kind)
  sm <- summary(m)

  if (kind %in% c("any","param")) {
    if (!is.null(sm$p.table) && nrow(sm$p.table) > 0) {
      pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)","Pr(>|z|)"))[1]
      if (!is.na(pcol) && var %in% rownames(sm$p.table)) {
        return(as.numeric(sm$p.table[var, pcol]))
      }
    }
  }

  if (kind %in% c("any","smooth")) {
    if (!is.null(sm$s.table) && nrow(sm$s.table) > 0) {
      sname <- paste0("s(", var, ")")
      if (sname %in% rownames(sm$s.table)) {
        return(as.numeric(sm$s.table[sname, "p-value"]))
      }
    }
  }

  NA_real_
}

# Plot a binary covariate effect (0 vs 1) with uncertainty, holding other fish at mean/zero/one
plot_binary_single_component <- function(model, var, data,
                                         family = c("binomial","gamma"),
                                         S = 2000, alpha = 0.05,
                                         exclude_RE = TRUE,
                                         hold_binaries = c("mean","zero","one"),
                                         id = "LID",
                                         fish_covars = c("Presence_Goldfish","Presence_Eel","Presence_Catfish","Presence_Common_smelt"),
                                         seed = 1) {
  family <- match.arg(family)
  hold_binaries <- match.arg(hold_binaries)

  # check term presence
  tl <- attr(terms(model), "term.labels")
  has_term <- any(tl == var) || any(grepl(paste0("^s\\(", var, "(,|\\))"), tl))
  if (!has_term) return(ggplot() + theme_void())

  # base at numeric means (+ default id level)
  num_means <- data %>%
    dplyr::summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
  base <- as.list(num_means)
  if (id %in% names(data) && is.factor(data[[id]]))
    base[[id]] <- factor(levels(data[[id]])[1], levels = levels(data[[id]]))

  # hold other fish binaries (plotting only)
  others <- setdiff(intersect(fish_covars, names(data)), var)
  set_bin <- function(x){
    if (hold_binaries == "mean") round(mean(data[[x]], na.rm = TRUE))
    else if (hold_binaries == "zero") 0L else 1L
  }
  for (x in others) base[[x]] <- set_bin(x)

  mk_nd <- function(level){
    nd <- as.data.frame(base, stringsAsFactors = FALSE)
    nd[[var]] <- as.integer(level)
    if (id %in% names(data) && is.factor(data[[id]])) {
      nd <- do.call(
        rbind,
        lapply(levels(data[[id]]), function(lv){
          r <- nd
          r[[id]] <- factor(lv, levels = levels(data[[id]]))
          r
        })
      )
    }
    if (!is.null(model$model)) {
      common <- intersect(names(nd), names(model$model))
      for (nm in common) if (is.factor(model$model[[nm]]))
        nd[[nm]] <- factor(nd[[nm]], levels = levels(model$model[[nm]]))
    }
    nd
  }
  nd0 <- mk_nd(0L)
  nd1 <- mk_nd(1L)
  nd  <- dplyr::bind_rows(
    dplyr::mutate(nd0, .level = 0L),
    dplyr::mutate(nd1, .level = 1L)
  )

  excl <- if (exclude_RE) exclude_if_RE(model) else NULL
  lp <- predict(model, newdata = nd, type = "link", se.fit = TRUE, exclude = excl)

  set.seed(seed)
  n <- nrow(nd)
  Z <- matrix(
    rnorm(n * S, lp$fit, pmax(lp$se.fit, .Machine$double.eps)),
    nrow = n, ncol = S
  )

  if (family == "binomial") {
    Y <- plogis(Z); ylab <- "Presence probability"
  } else {
    Y <- exp(Z);    ylab <- "Mean given presence (μ)"
  }

  sim_by_level <- function(level_flag){
    sims <- Y[nd$.level == level_flag, , drop = FALSE]
    colMeans(sims)
  }
  sim0 <- sim_by_level(0L)
  sim1 <- sim_by_level(1L)

  qlo <- alpha/2
  qhi <- 1 - alpha/2
  df <- data.frame(
    level = factor(c(0,1), levels = c(0,1), labels = c("Absent","Present")),
    mean  = c(mean(sim0), mean(sim1)),
    lwr   = c(quantile(sim0, qlo), quantile(sim1, qlo)),
    upr   = c(quantile(sim0, qhi), quantile(sim1, qhi))
  )

  ggplot(df, aes(x = level, y = mean)) +
    geom_col(width = 0.6) +
    geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2) +
    labs(x = var, y = ylab)
}

as_plot <- function(p) if (inherits(p, c("gg","ggplot","patchwork"))) p else patchwork::plot_spacer()

# DATA PREP HELPERS

prepare_block <- function(Modeling_data, response, vars_common, id = "LID"){
  vars <- c(response, vars_common)
  out <- Modeling_data %>%
    dplyr::select(dplyr::all_of(vars)) %>%
    dplyr::mutate(
      "{id}" := factor(.data[[id]]),                        # LID as factor (lake)
      dplyr::across(dplyr::any_of(fish_vars), ~ as.numeric(.x))
    )
  out
}

# Row of neutral reference values
ref_row <- function(data){
  dplyr::summarise(
    data,
    dplyr::across(
      dplyr::everything(),
      \(x){
        if (is.numeric(x)) stats::median(x, na.rm = TRUE)
        else if (is.factor(x)) levels(x)[1L]
        else if (is.logical(x)) FALSE
        else if (is.character(x)) unique(stats::na.omit(x))[1L]
        else x[1L]
      }
    )
  )
}

is_binary_numeric <- function(x) is.numeric(x) && dplyr::n_distinct(x, na.rm = TRUE) == 2

# Get non-smooth model terms that exist in data (parametric)
get_parametric_terms <- function(model, data){
  tl <- attr(stats::terms(model), "term.labels") # e.g. "Presence_Eel", "s(pH)"
  param_labels <- tl[!grepl("^(s|te|ti|t2)\\(", tl)]   # drop s(), te(), ti(), t2()
  param_labels[param_labels %in% names(data)]
}

# Plot parametric numeric (continuous) partial on response scale
plot_param_numeric <- function(model, var, data, n = 100){
  r <- range(data[[var]], na.rm = TRUE)
  grid <- data.frame(x = seq(r[1], r[2], length.out = n))
  names(grid) <- var

  base <- ref_row(data)
  newd <- base[rep(1, nrow(grid)), , drop = FALSE]
  newd[[var]] <- grid[[var]]

  pr  <- predict(model, newdata = newd, se.fit = TRUE, type = "link")
  il  <- model$family$linkinv
  fit <- il(pr$fit)
  lo  <- il(pr$fit - 1.96*pr$se.fit)
  hi  <- il(pr$fit + 1.96*pr$se.fit)

  ggplot2::ggplot(
    data.frame(x = newd[[var]], fit, lo, hi),
    ggplot2::aes(x = .data[[var]], y = fit)
  ) +
    ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.2) +
    ggplot2::geom_line() +
    ggplot2::labs(x = var, y = "Partial effect (response scale)")
}

# Plot parametric binary/factor partial on response scale
plot_param_factor <- function(model, var, data){
  mod_is_factor  <- !is.null(model$model) && is.factor(model$model[[var]])
  mod_is_numeric <- !is.null(model$model) && is.numeric(model$model[[var]])

  if (mod_is_numeric || is_binary_numeric(data[[var]])) {
    lev_vals <- c(0, 1)
    lev_labs <- c("0","1")
  } else if (is.factor(data[[var]])) {
    lev_vals <- levels(data[[var]])
    lev_labs <- lev_vals
  } else {
    lev_vals <- sort(unique(stats::na.omit(data[[var]])))
    lev_labs <- as.character(lev_vals)
  }

  base <- ref_row(data)
  newd <- base[rep(1, length(lev_vals)), , drop = FALSE]

  if (mod_is_numeric) {
    newd[[var]] <- as.numeric(lev_vals)
  } else {
    newd[[var]] <- factor(lev_vals, levels = lev_vals)
  }

  if (!is.null(model$model)) {
    common <- intersect(names(newd), names(model$model))
    for (nm in common) if (is.factor(model$model[[nm]]))
      newd[[nm]] <- factor(newd[[nm]], levels = levels(model$model[[nm]]))
  }

  pr  <- predict(model, newdata = newd, se.fit = TRUE, type = "link")
  il  <- model$family$linkinv
  fit <- il(pr$fit)
  lo  <- il(pr$fit - 1.96*pr$se.fit)
  hi  <- il(pr$fit + 1.96*pr$se.fit)

  x_for_plot <- factor(lev_labs, levels = lev_labs)

  ggplot2::ggplot(
    data.frame(x = x_for_plot, fit, lo, hi),
    ggplot2::aes(x = x, y = fit)
  ) +
    ggplot2::geom_point() +
    ggplot2::geom_errorbar(
      ggplot2::aes(ymin = lo, ymax = hi),
      width = 0.15
    ) +
    ggplot2::labs(x = var, y = "Partial effect (response scale)")
}

# Dispatch: prefer gratia for smooths, otherwise parametric plot (factor vs numeric)
plot_single_component <- function(model, var, data){
  sm_lab <- paste0("s(", var, ")")
  if (sm_lab %in% gratia::smooths(model)) {
    return(gratia::draw(model, select = sm_lab))
  }
  if (is.factor(data[[var]]) || is_binary_numeric(data[[var]])) {
    return(plot_param_factor(model, var, data))
  }
  plot_param_numeric(model, var, data)
}

# Significant parametric terms (LID normally excluded here because it is a random effect)
significant_parametric_terms <- function(model, data, alpha = 0.1, exclude = c("LID")){
  params <- setdiff(get_parametric_terms(model, data), exclude)
  keep <- vapply(params, \(v){
    p <- get_term_p(model, v)
    !is.na(p) && p <= alpha
  }, logical(1))
  params[keep]
}

plot_linear_single_component <- function(model, var, data,
                                         family = c("binomial","gamma"),
                                         n = 100, alpha = 0.05,
                                         exclude_RE = TRUE) {
  family <- match.arg(family)

  num_means <- data %>%
    dplyr::summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
  base <- as.list(num_means)

  if (!is.null(model$model)) {
    for (nm in names(model$model)) {
      if (is.factor(model$model[[nm]])) {
        base[[nm]] <- factor(
          levels(model$model[[nm]])[1],
          levels = levels(model$model[[nm]])
        )
      }
    }
  }

  x  <- data[[var]]
  xr <- range(x, na.rm = TRUE)
  grid <- seq(xr[1], xr[2], length.out = n)

  nd <- as.data.frame(base)
  nd <- nd[rep(1, n), , drop = FALSE]
  nd[[var]] <- grid

  excl <- if (exclude_RE) exclude_if_RE(model) else NULL
  pr <- predict(model, newdata = nd, type = "link", se.fit = TRUE, exclude = excl)

  if (family == "binomial") {
    fit <- plogis(pr$fit)
    lwr <- plogis(pr$fit - qnorm(1 - alpha/2) * pr$se.fit)
    upr <- plogis(pr$fit + qnorm(1 - alpha/2) * pr$se.fit)
    ylab <- "Response (probability)"
  } else {
    fit <- exp(pr$fit)
    lwr <- exp(pr$fit - qnorm(1 - alpha/2) * pr$se.fit)
    upr <- exp(pr$fit + qnorm(1 - alpha/2) * pr$se.fit)
    ylab <- "Response (μ)"
  }

  df <- data.frame(x = grid, fit = fit, lwr = lwr, upr = upr)
  ggplot(df, aes(x, fit)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
    geom_line() +
    labs(x = var, y = ylab)
}

# Build a panel of parametric plots (significant-only by default)
parametric_panel <- function(model, data, alpha = 0.1, exclude = character(0),
                             family = c("binomial","gamma"),
                             fish_covars = c("Presence_Goldfish","Presence_Eel","Presence_Catfish","Presence_Common_smelt"),
                             hold_binaries = "mean") {
  family <- match.arg(family)

  sm <- summary(model)
  if (is.null(sm$p.table) || nrow(sm$p.table) == 0) return(ggplot() + theme_void())

  pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)","Pr(>|z|)"))[1]
  if (is.na(pcol)) return(ggplot() + theme_void())

  par_names <- setdiff(rownames(sm$p.table), "(Intercept)")

  smooth_rows <- if (!is.null(sm$s.table)) rownames(sm$s.table) else character(0)
  smooth_vars <- sub("^s\\(([^,\\)]+).*$", "\\1", smooth_rows)
  par_names <- setdiff(par_names, union(exclude, smooth_vars))

  par_sig <- par_names[sm$p.table[par_names, pcol, drop = TRUE] <= alpha]
  if (!length(par_sig)) return(ggplot() + theme_void())

  plots <- lapply(par_sig, function(v) {
    x <- data[[v]]
    if (is.null(x)) return(ggplot() + theme_void())

    is_bin_num <- is.numeric(x) && dplyr::n_distinct(x, na.rm = TRUE) <= 2
    is_bin_fac <- is.factor(x)  && nlevels(x) == 2

    if (is_bin_num || is_bin_fac) {
      plot_binary_single_component(
        model, v, data,
        family        = family,
        hold_binaries = hold_binaries,
        fish_covars   = fish_covars
      )
    } else {
      plot_linear_single_component(model, v, data, family = family)
    }
  })

  if (!length(plots)) return(ggplot() + theme_void())
  patchwork::wrap_plots(plots, ncol = 1)
}
```

# OCCUPANCY: Presence/Absence (binomial analysis)
```{r fig.width=14, fig.height=8, dpi=100, echo=FALSE, message=FALSE, warning=FALSE}
PData <- prepare_block(Modeling_data, "Presence_Kōura", vars_common)

# 2) Candidate predictors for VIF (fixed effects only: exclude response and LID)
id <- "LID"
pred_fixed_occ <- setdiff(names(PData), c("Presence_Kōura", id))

# 3) VIF pruning (binomial) – all fixed effects treated the same
vif_occ <- remove_high_vif_glmBI(
  PData,
  "Presence_Kōura",
  pred_fixed_occ,
  threshold   = vif_thresh,
  protect_vars = character(0))
kept_fixed_occ <- vif_occ$predictors

vif_occ$report$start
vif_occ$report$removed
vif_occ$report$final_vif

# 4) Build FULL (ML) formula: fixed effects + LID random effect (if INCLUDE_RE)
vars_step_occ <- c(kept_fixed_occ, if (INCLUDE_RE) id)
rhs_full_occ <- paste(
  vapply(
    vars_step_occ,
    function(v) build_smooth_gam(v, PData, custom_k, include_re_for_LID = INCLUDE_RE),
    character(1)),
  collapse = " + ")
form_full_occ <- as.formula(paste("Presence_Kōura ~", rhs_full_occ))

# 5) Fit FULL (ML)
full_ml_occ <- mgcv::gam(
  form_full_occ,
  data   = PData,
  family = binomial(link = "logit"),
  method = "ML",
  select = TRUE)

# 6) Stepwise (ML) by p-values, with LID random effect protected
remaining_occ <- vars_step_occ
protected_step_vars <- if (INCLUDE_RE) id else character(0)

repeat {
  # fit current model
  rhs_now <- paste(
    vapply(
      remaining_occ,
      function(v) build_smooth_gam(v, PData, custom_k, include_re_for_LID = INCLUDE_RE),
      character(1)
    ),
    collapse = " + "
  )
  m_now <- mgcv::gam(
    as.formula(paste("Presence_Kōura ~", rhs_now)),
    data   = PData,
    family = binomial(link = "logit"),
    method = "ML",
    select = TRUE
  )
  sm <- summary(m_now)

  # collect p-values for parametric & smooth terms
  ps <- c()
  if (!is.null(sm$p.table) && nrow(sm$p.table) > 0) {
    pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)","Pr(>|z|)"))[1]
    pvec <- sm$p.table[, pcol]
    ps <- c(ps, pvec[names(pvec) != "(Intercept)"])
  }
  if (!is.null(sm$s.table) && nrow(sm$s.table) > 0) {
    pvec <- sm$s.table[, "p-value"]
    names(pvec) <- rownames(sm$s.table)
    ps <- c(ps, pvec)
  }
  drop_candidates <- ps[ps > p_cutoff_ml]

  if (!length(drop_candidates)) { red_ml_occ <- m_now; break }

  ordered <- names(sort(drop_candidates, decreasing = TRUE))
  # translate e.g. "s(pH)" -> "pH"
  ordered_vars <- vapply(
    ordered,
    function(x) if (grepl("^s\\(", x)) sub("^s\\(([^,]+).*\\)$", "\\1", x) else x,
    character(1)
  )

  # never drop protected variables (here: LID when INCLUDE_RE = TRUE)
  ordered_vars <- setdiff(ordered_vars, protected_step_vars)
  if (!length(ordered_vars)) { red_ml_occ <- m_now; break }

  remove_v <- ordered_vars[1]
  message("Dropping variable: ", remove_v)

  remaining_occ <- setdiff(remaining_occ, remove_v)
  if (!length(remaining_occ)) { red_ml_occ <- m_now; break }
}

# ML comparison: FULL vs REDUCED
summary(full_ml_occ)
summary(red_ml_occ)
AIC(full_ml_occ, red_ml_occ)
suppressWarnings(print(anova(full_ml_occ, red_ml_occ, test = "Chisq")))

# Clean test of the LID random effect: same fixed structure, with vs without s(LID, bs='re')
LID_test_occ <- if (INCLUDE_RE) {
  # formula for reduced ML model with LID random effect
  f_withLID <- formula(red_ml_occ)
  # remove the LID random effect term only (keep fixed effects identical)
  f_noLID <- update(f_withLID, . ~ . - s(LID, bs = "re"))
  m_withLID <- red_ml_occ
  m_noLID   <- mgcv::gam(
    f_noLID,
    data   = PData,
    family = binomial(link = "logit"),
    method = "ML",
    select = TRUE)
  anova(m_noLID, m_withLID, test = "Chisq")
}

# 7) Final REML fit using the selected structure (with LID RE if INCLUDE_RE = TRUE)
final_occ <- mgcv::gam(
  formula(red_ml_occ),
  data   = PData,
  family = binomial(link = "logit"),
  method = "REML",
  select = TRUE)

summary(final_occ)
mgcv::concurvity(final_occ)
gam.check(final_occ)

# 8) Plots (Hydrobiologia-compliant styling)
base_theme_bw <- theme_classic() +
  theme(
    text = element_text(family = "Arial", size = 8),
    axis.title = element_text(face = "plain"),
    axis.text = element_text(face = "plain"),
    plot.title = element_text(face = "plain"),
    strip.text = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3)
  )

clean_smooth_title <- function(x) {
  x <- gsub("^s\\((.*)\\)$", "\\1", x)
  x <- gsub("_", " ", x)
  x
}

# Smooth partial effects (NO shading, NO rug)
# NOTE: p_occ_smooths is now built as 4 individual smooth panels for full control
sm <- smooth_estimates(final_occ)
smooth_names <- unique(sm$.smooth)
smooth_4_names <- smooth_names[1:4]

plots_4 <- lapply(smooth_4_names, function(s) {
  gratia::draw(final_occ, select = s, scales = "fixed", se = FALSE) +
    labs(title = clean_smooth_title(s), x = NULL, y = "Partial effect") +
    base_theme_bw
})

final_p_occ_smooths <- patchwork::wrap_plots(plots_4, ncol = 4)

# Parametric terms panel (kept; style enforced)
p_occ_param <- parametric_panel(
  final_occ,
  PData,
  alpha       = .1,
  exclude     = c("LID"),
  family      = "binomial",
  fish_covars = fish_vars
) + base_theme_bw

# ROC curve + AUC (population-level prediction, excluding LID RE)
pred_pres <- predict(
  final_occ, type    = "response",
  exclude = exclude_if_RE(final_occ))

obs_pres <- PData$Presence_Kōura == 1
roc_obj  <- pROC::roc(obs_pres, pred_pres)
auc_val  <- as.numeric(pROC::auc(roc_obj))

roc_df <- data.frame(
  tpr = roc_obj$sensitivities,
  fpr = 1 - roc_obj$specificities)

p_pres_roc <- ggplot(roc_df, aes(fpr, tpr)) +
  geom_path(colour = "black", linewidth = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black", linewidth = 0.4) +
  annotate("text", x = 0.65, y = 0.1, label = paste0("AUC = ", round(auc_val, 3)), size = 3) +
  labs(x = "False positive rate", y = "True positive rate") +
  base_theme_bw

# Simple grouped CV-by-lake for calibration (LID used only as a grouping factor here)
set.seed(1)
k_occ       <- min(5, nlevels(PData$LID))
lid_levels  <- levels(PData$LID)
lid_folds   <- sample(rep(1:k_occ, length.out = length(lid_levels)))
lid2fold    <- setNames(lid_folds, lid_levels)
fold_vec    <- unname(lid2fold[as.character(PData$LID)])
pred_cv_occ <- rep(NA_real_, nrow(PData))

# Terms present in the final model (strip s(...))
final_terms <- {
  tl <- attr(terms(final_occ), "term.labels")
  sub("^s\\(([^,\\)]+).*$", "\\1", tl)
}

for (fold in 1:k_occ) {
  idx   <- which(fold_vec == fold)
  train <- PData[-idx, , drop = FALSE]
  test  <- PData[idx,  , drop = FALSE]

  rhs_now <- paste(
    vapply(
      final_terms,
      function(v) build_smooth_gam(v, train, custom_k, include_re_for_LID = INCLUDE_RE),
      character(1)
    ),
    collapse = " + "
  )

  fml_now <- as.formula(paste("Presence_Kōura ~", rhs_now))

  m <- mgcv::gam(
    fml_now,
    data   = train,
    family = binomial(link = "logit"),
    method = "REML",
    select = TRUE
  )

  for (nm in names(test)) {
    if (is.factor(train[[nm]])) {
      test[[nm]] <- factor(test[[nm]], levels = levels(train[[nm]]))
    }
  }

  pred_cv_occ[idx] <- predict(
    m,
    newdata = test,
    type    = "response",
    exclude = exclude_if_RE(m)
  )
}

calib_data_occ <- tibble(pred = pred_cv_occ, obs = PData$Presence_Kōura) %>%
  mutate(bin = ntile(pred, 5)) %>%
  group_by(bin) %>%
  summarise(
    mean_pred = mean(pred, na.rm = TRUE),
    obs_rate  = mean(obs,  na.rm = TRUE),
    n         = dplyr::n(),
    .groups   = "drop"
  )

Calibration_plot_occ <- ggplot(calib_data_occ, aes(mean_pred, obs_rate)) +
  geom_point(size = 2.2, colour = "black") +
  geom_line(colour = "black", linewidth = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black", linewidth = 0.4) +
  labs(x = "Mean predicted probability", y = "Observed proportion") +
  base_theme_bw

# Layout: top row (smooths), bottom row (ROC + calibration)
top_row <- final_p_occ_smooths + patchwork::plot_layout(ncol = 4)
bottom_row <- (p_pres_roc + Calibration_plot_occ) + patchwork::plot_layout(ncol = 2)

final_plot_occupancy <- (top_row / bottom_row) +
  patchwork::plot_layout(heights = c(1, 2))

final_plot_occupancy

ggsave(filename = file.path(fig_dir, "Occupancy_panels.png"), plot = final_plot_occupancy, width = 180, height = 120, units = "mm", dpi = 1200)

```

# CPUE: HURDLE
```{r fig.width=10, fig.height=5, dpi=100, echo=FALSE, message=FALSE, warning=FALSE}
CData <- prepare_block(Modeling_data, "Weighted_CPUE_Kōura", vars_common) %>%
  dplyr::mutate(CPUE_pos = as.integer(Weighted_CPUE_Kōura > 0))

id <- "LID"

# ---------- Presence component (binomial)

# VIF candidates (exclude response, id, and raw CPUE columns)
pred_fixed_cpue_pres <- setdiff(
  names(CData),
  c("CPUE_pos", id, "Weighted_CPUE_Kōura", "Weighted_CPUE_Koura", "CPUE"))

# VIF pruning: all fixed effects treated the same
vif_cpue_pres <- remove_high_vif_glmBI(
  CData,
  "CPUE_pos",
  pred_fixed_cpue_pres,
  threshold    = vif_thresh,
  protect_vars = character(0))
kept_fixed_cpue_pres <- vif_cpue_pres$predictors

vif_cpue_pres$report$start
vif_cpue_pres$report$removed
vif_cpue_pres$report$final_vif


# Stepwise set: fixed effects + LID random effect (if requested)
vars_step_cpue_pres <- c(kept_fixed_cpue_pres, if (INCLUDE_RE) id)

rhs_cpue_pres <- paste(
  vapply(
    vars_step_cpue_pres,
    function(v) build_smooth_gam(v, CData, custom_k, include_re_for_LID = INCLUDE_RE),
    character(1)),collapse = " + ")

full_ml_cpue_pres <- mgcv::gam(
  as.formula(paste("CPUE_pos ~", rhs_cpue_pres)),
  data   = CData,
  family = binomial(link = "logit"),
  method = "ML",
  select = TRUE)

# Stepwise (ML) with LID protected if used as random effect
remaining_cpue_pres   <- vars_step_cpue_pres
protected_step_vars   <- if (INCLUDE_RE) id else character(0)

repeat {
  rhs_now <- paste(
    vapply(
      remaining_cpue_pres,
      function(v) build_smooth_gam(v, CData, custom_k, include_re_for_LID = INCLUDE_RE),
      character(1)
    ),
    collapse = " + "
  )

  m_now <- mgcv::gam(
    as.formula(paste("CPUE_pos ~", rhs_now)),
    data   = CData,
    family = binomial(link = "logit"),
    method = "ML",
    select = TRUE
  )
  sm <- summary(m_now)

  ps <- c()
  if (!is.null(sm$p.table) && nrow(sm$p.table) > 0) {
    pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)", "Pr(>|z|)"))[1]
    pvec <- sm$p.table[, pcol]
    ps   <- c(ps, pvec[names(pvec) != "(Intercept)"])
  }
  if (!is.null(sm$s.table) && nrow(sm$s.table) > 0) {
    pvec <- sm$s.table[, "p-value"]
    names(pvec) <- rownames(sm$s.table)
    ps   <- c(ps, pvec)
  }
  drop_candidates <- ps[ps > p_cutoff_ml]
  if (!length(drop_candidates)) { red_ml_cpue_pres <- m_now; break }

  ordered <- names(sort(drop_candidates, decreasing = TRUE))
  ordered_vars <- vapply(
    ordered,
    function(x) if (grepl("^s\\(", x))
      sub("^s\\(([^,]+).*\\)$", "\\1", x) else x,
    character(1)
  )

  # never drop protected variables (LID as random effect)
  ordered_vars <- setdiff(ordered_vars, protected_step_vars)
  if (!length(ordered_vars)) { red_ml_cpue_pres <- m_now; break }

  remove_v <- ordered_vars[1]
  remaining_cpue_pres <- setdiff(remaining_cpue_pres, remove_v)
  if (!length(remaining_cpue_pres)) { red_ml_cpue_pres <- m_now; break }
}

# ML comparison: FULL vs REDUCED 
summary(full_ml_cpue_pres)
summary(red_ml_cpue_pres)
print(AIC(full_ml_cpue_pres, red_ml_cpue_pres))
suppressWarnings(print(anova(full_ml_cpue_pres, red_ml_cpue_pres, test = "Chisq")))

# Clean test of LID random effect: same fixed effects, with vs without s(LID, bs='re')
if (INCLUDE_RE) {
  # current reduced ML model (with LID random effect)
  m_withLID <- red_ml_cpue_pres

  # terms in the model
  tl <- attr(terms(red_ml_cpue_pres), "term.labels")
  # remove any term involving LID
  tl_noLID <- tl[!grepl("LID", tl)]
  rhs_noLID <- if (length(tl_noLID)) paste(tl_noLID, collapse = " + ") else "1"
  f_noLID <- as.formula(paste("CPUE_pos ~", rhs_noLID))

  m_noLID <- mgcv::gam(
    f_noLID,
    data   = CData,
    family = binomial(link = "logit"),
    method = "ML",
    select = TRUE
  )

  anova(m_noLID, m_withLID, test = "Chisq")
}

# Final REML
final_cpue_pres <- mgcv::gam(
  formula(red_ml_cpue_pres),
  data   = CData,
  family = binomial(link = "logit"),
  method = "REML",
  select = TRUE
)
summary(final_cpue_pres)
mgcv::concurvity(final_cpue_pres)
gam.check(final_cpue_pres)

# Plots (presence CPUE)
p_cpue_pres_smooths <- gratia::draw(final_cpue_pres)
p_cpue_pres_param <- parametric_panel(
  final_cpue_pres,
  CData,
  alpha       = alpha_sig,
  exclude     = c("LID"),
  family      = "binomial",
  fish_covars = fish_vars
)

# ROC (presence CPUE, population-level predictions)
pred_pres <- predict(
  final_cpue_pres,
  type    = "response",
  exclude = exclude_if_RE(final_cpue_pres)
)
obs_pres  <- CData$CPUE_pos
roc_obj   <- pROC::roc(obs_pres, pred_pres)
auc_val   <- as.numeric(pROC::auc(roc_obj))
roc_df    <- data.frame(
  tpr = roc_obj$sensitivities,
  fpr = 1 - roc_obj$specificities
)
p_cpue_pres_roc <- ggplot(roc_df, aes(fpr, tpr)) +
  geom_path() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  annotate("text", x = 0.6, y = 0.1,
           label = paste0("AUC = ", round(auc_val, 3))) +
  labs(x = "False positive rate", y = "True positive rate")

CPUE_panel_presence <- patchwork::wrap_plots(
  list(
    as_plot(p_cpue_pres_smooths),
    as_plot(p_cpue_pres_param),
    as_plot(p_cpue_pres_roc)
  ),
  ncol = 3
)
CPUE_panel_presence

# ---------- Positives component (Gamma, log link)

Cpos <- dplyr::filter(CData, CPUE_pos == 1L)

pred_fixed_cpue_pos <- setdiff(names(Cpos), c("Weighted_CPUE_Kōura", id))

vif_cpue_pos <- remove_high_vif_glmmTMB(
  Cpos,
  "Weighted_CPUE_Kōura",
  pred_fixed_cpue_pos,
  threshold    = vif_thresh,
  protect_vars = character(0)
)
kept_fixed_cpue_pos <- vif_cpue_pos$predictors

vif_cpue_pos$report$start
vif_cpue_pos$report$removed
vif_cpue_pos$report$final_vif


vars_step_cpue_pos <- c(kept_fixed_cpue_pos, if (INCLUDE_RE) id)

rhs_cpue_pos <- paste(
  vapply(
    vars_step_cpue_pos,
    function(v) build_smooth_gam(v, Cpos, custom_k, include_re_for_LID = INCLUDE_RE),
    character(1)
  ),
  collapse = " + "
)

full_ml_cpue_pos <- mgcv::gam(
  as.formula(paste("Weighted_CPUE_Kōura ~", rhs_cpue_pos)),
  data   = Cpos,
  family = Gamma(link = "log"),
  method = "ML",
  select = TRUE
)

remaining_cpue_pos <- vars_step_cpue_pos
protected_step_vars_pos <- if (INCLUDE_RE) id else character(0)

repeat {
  rhs_now <- paste(
    vapply(
      remaining_cpue_pos,
      function(v) build_smooth_gam(v, Cpos, custom_k, include_re_for_LID = INCLUDE_RE),
      character(1)
    ),
    collapse = " + "
  )
  m_now <- mgcv::gam(
    as.formula(paste("Weighted_CPUE_Kōura ~", rhs_now)),
    data   = Cpos,
    family = Gamma(link = "log"),
    method = "ML",
    select = TRUE
  )
  sm <- summary(m_now)

  ps <- c()
  if (!is.null(sm$p.table) && nrow(sm$p.table) > 0) {
    pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)", "Pr(>|z|)"))[1]
    pvec <- sm$p.table[, pcol]
    ps   <- c(ps, pvec[names(pvec) != "(Intercept)"])
  }
  if (!is.null(sm$s.table) && nrow(sm$s.table) > 0) {
    pvec <- sm$s.table[, "p-value"]
    names(pvec) <- rownames(sm$s.table)
    ps   <- c(ps, pvec)
  }
  drop_candidates <- ps[ps > p_cutoff_ml]
  if (!length(drop_candidates)) { red_ml_cpue_pos <- m_now; break }

  ordered <- names(sort(drop_candidates, decreasing = TRUE))
  ordered_vars <- vapply(
    ordered,
    function(x) if (grepl("^s\\(", x))
      sub("^s\\(([^,]+).*\\)$", "\\1", x) else x,
    character(1)
  )

  # protect LID in the positive model as well
  ordered_vars <- setdiff(ordered_vars, protected_step_vars_pos)
  if (!length(ordered_vars)) { red_ml_cpue_pos <- m_now; break }

  remove_v <- ordered_vars[1]
  remaining_cpue_pos <- setdiff(remaining_cpue_pos, remove_v)
  if (!length(remaining_cpue_pos)) { red_ml_cpue_pos <- m_now; break }
}

# ML comparison: FULL vs REDUCED
summary(full_ml_cpue_pos)
summary(red_ml_cpue_pos)
print(AIC(full_ml_cpue_pos, red_ml_cpue_pos))
suppressWarnings(print(anova(full_ml_cpue_pos, red_ml_cpue_pos, test = "Chisq")))

# Clean test of LID random effect for positive CPUE
LID_test_cpue <- if (INCLUDE_RE) {
  m_withLID_pos <- red_ml_cpue_pos
  tl_pos <- attr(terms(red_ml_cpue_pos), "term.labels")
  tl_pos_noLID <- tl_pos[!grepl("LID", tl_pos)]
  rhs_pos_noLID <- if (length(tl_pos_noLID)) paste(tl_pos_noLID, collapse = " + ") else "1"
  f_pos_noLID <- as.formula(paste("Weighted_CPUE_Kōura ~", rhs_pos_noLID))

  m_noLID_pos <- mgcv::gam(
    f_pos_noLID,
    data   = Cpos,
    family = Gamma(link = "log"),
    method = "ML",
    select = TRUE
  )

  anova(m_noLID_pos, m_withLID_pos, test = "Chisq")
}

# Final REML
final_cpue_pos <- mgcv::gam(
  formula(red_ml_cpue_pos),
  data   = Cpos,
  family = Gamma(link = "log"),
  method = "REML",
  select = TRUE
)
summary(final_cpue_pos)
mgcv::concurvity(final_cpue_pos)
gam.check(final_cpue_pos)

# Plots (positive CPUE) – updated layout, no LID
base_theme_bw <- theme_classic() +
  theme(
    text = element_text(family = "Arial", size = 8),
    axis.title = element_text(face = "plain"),
    axis.text = element_text(face = "plain"),
    plot.title = element_text(face = "plain"),
    strip.text = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3)
  )

clean_smooth_title <- function(x) {
  x <- gsub("^s\\((.*)\\)$", "\\1", x)
  x <- gsub("_", " ", x)
  x
}

# Smooth partial effects (exclude LID explicitly)
sm_cpue <- smooth_estimates(final_cpue_pos)
cpue_smooth_names <- setdiff(unique(sm_cpue$.smooth), "s(LID)")
cpue_smooth_plots <- lapply(cpue_smooth_names, function(s) {
  gratia::draw(final_cpue_pos, select = s, scales = "fixed", se = FALSE) +
    labs(title = clean_smooth_title(s), x = NULL, y = "Partial effect") +
    base_theme_bw
})

p_cpue_pos_smooths <- patchwork::wrap_plots(cpue_smooth_plots, ncol = length(cpue_smooth_plots)) +
  plot_annotation(tag_levels = "a")

# Scatter diagnostics (population-level prediction, no LID)
pred_pos <- predict(final_cpue_pos, type = "response")
obs_pos  <- final_cpue_pos$y
R2   <- cor(pred_pos, obs_pos, use = "complete.obs")^2
RMSE <- sqrt(mean((pred_pos - obs_pos)^2, na.rm = TRUE))

df_scat <- data.frame(Predicted = pred_pos, Observed = obs_pos)

p_cpue_pos_scatter <- ggplot(df_scat, aes(Predicted, Observed)) +
  geom_point(size = 1.8, colour = "black") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black", linewidth = 0.4) +
  annotate(
    "text",
    x = max(df_scat$Predicted, na.rm = TRUE) * 0.7,
    y = max(df_scat$Observed,  na.rm = TRUE) * 0.8,
    label = paste0("R\u00B2 = ", round(R2, 2), "\nRMSE = ", round(RMSE, 2)),
    size = 3
  ) +
  labs(x = "Predicted CPUE", y = "Observed CPUE", tag = "b") +
  base_theme_bw

# Layout: flatter top, taller bottom
CPUE_panel_positives <- (p_cpue_pos_smooths / p_cpue_pos_scatter) +
  patchwork::plot_layout(heights = c(1, 2))

CPUE_panel_positives

ggsave( filename = file.path(fig_dir, "CPUE_panel_positives.png"), plot = CPUE_panel_positives, width = 130, height = 120, units = "mm", dpi = 1200)

```

# BPUE: HURDLE
```{r fig.width=10, fig.height=5, dpi=100, echo=FALSE, message=FALSE, warning=FALSE}
BData <- prepare_block(Modeling_data, "Weighted_BPUE_Kōura", vars_common) %>%
  dplyr::mutate(BPUE_pos = as.integer(Weighted_BPUE_Kōura > 0))

id <- "LID"

# ---------- Presence component (binomial)

pred_fixed_BPUE_pres <- setdiff(
  names(BData),
  c("BPUE_pos", id, "Weighted_BPUE_Kōura", "Weighted_BPUE_Koura", "Weighted_BPUE"))

vif_BPUE_pres <- remove_high_vif_glmBI(
  BData,
  "BPUE_pos",
  pred_fixed_BPUE_pres,
  threshold    = vif_thresh,
  protect_vars = character(0))
kept_fixed_BPUE_pres <- vif_BPUE_pres$predictors

vif_BPUE_pres$report$start
vif_BPUE_pres$report$removed
vif_BPUE_pres$report$final_vif

# Stepwise set: fixed effects + LID random effect (if requested)
vars_step_BPUE_pres <- c(kept_fixed_BPUE_pres, if (INCLUDE_RE) id)

rhs_BPUE_pres <- paste(
  vapply(
    vars_step_BPUE_pres,
    function(v) build_smooth_gam(v, BData, custom_k, include_re_for_LID = INCLUDE_RE),
    character(1)
  ),
  collapse = " + "
)

full_ml_BPUE_pres <- mgcv::gam(
  as.formula(paste("BPUE_pos ~", rhs_BPUE_pres)),
  data   = BData,
  family = binomial(link = "logit"),
  method = "ML",
  select = TRUE
)

remaining_BPUE_pres <- vars_step_BPUE_pres
protected_step_vars_b <- if (INCLUDE_RE) id else character(0)

repeat {
  rhs_now <- paste(
    vapply(
      remaining_BPUE_pres,
      function(v) build_smooth_gam(v, BData, custom_k, include_re_for_LID = INCLUDE_RE),
      character(1)
    ),
    collapse = " + "
  )
  m_now <- mgcv::gam(
    as.formula(paste("BPUE_pos ~", rhs_now)),
    data   = BData,
    family = binomial(link = "logit"),
    method = "ML",
    select = TRUE
  )
  sm <- summary(m_now)

  ps <- c()
  if (!is.null(sm$p.table) && nrow(sm$p.table) > 0) {
    pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)", "Pr(>|z|)"))[1]
    pvec <- sm$p.table[, pcol]
    ps   <- c(ps, pvec[names(pvec) != "(Intercept)"])
  }
  if (!is.null(sm$s.table) && nrow(sm$s.table) > 0) {
    pvec <- sm$s.table[, "p-value"]
    names(pvec) <- rownames(sm$s.table)
    ps   <- c(ps, pvec)
  }
  drop_candidates <- ps[ps > p_cutoff_ml]
  if (!length(drop_candidates)) { red_ml_BPUE_pres <- m_now; break }

  ordered <- names(sort(drop_candidates, decreasing = TRUE))
  ordered_vars <- vapply(
    ordered,
    function(x) if (grepl("^s\\(", x))
      sub("^s\\(([^,]+).*\\)$", "\\1", x) else x,
    character(1)
  )

  # protect LID
  ordered_vars <- setdiff(ordered_vars, protected_step_vars_b)
  if (!length(ordered_vars)) { red_ml_BPUE_pres <- m_now; break }

  remove_v <- ordered_vars[1]
  remaining_BPUE_pres <- setdiff(remaining_BPUE_pres, remove_v)
  if (!length(remaining_BPUE_pres)) { red_ml_BPUE_pres <- m_now; break }
}

# ML comparison: FULL vs REDUCED 
summary(full_ml_BPUE_pres)
summary(red_ml_BPUE_pres)
print(AIC(full_ml_BPUE_pres, red_ml_BPUE_pres))
suppressWarnings(print(anova(full_ml_BPUE_pres, red_ml_BPUE_pres, test = "Chisq")))

# Clean test of LID random effect for BPUE presence
if (INCLUDE_RE) {
  m_withLID_b <- red_ml_BPUE_pres
  tl_b <- attr(terms(red_ml_BPUE_pres), "term.labels")
  tl_b_noLID <- tl_b[!grepl("LID", tl_b)]
  rhs_b_noLID <- if (length(tl_b_noLID)) paste(tl_b_noLID, collapse = " + ") else "1"
  f_b_noLID <- as.formula(paste("BPUE_pos ~", rhs_b_noLID))

  m_noLID_b <- mgcv::gam(
    f_b_noLID,
    data   = BData,
    family = binomial(link = "logit"),
    method = "ML",
    select = TRUE
  )

  anova(m_noLID_b, m_withLID_b, test = "Chisq")
}

# Final REML
final_BPUE_pres <- mgcv::gam(
  formula(red_ml_BPUE_pres),
  data   = BData,
  family = binomial(link = "logit"),
  method = "REML",
  select = TRUE
)
summary(final_BPUE_pres)
mgcv::concurvity(final_BPUE_pres)
gam.check(final_BPUE_pres)

# Plots (BPUE presence)
p_BPUE_pres_smooths <- gratia::draw(final_BPUE_pres)
p_BPUE_pres_param <- parametric_panel(
  final_BPUE_pres,
  BData,
  alpha       = alpha_sig,
  exclude     = c("LID"),
  family      = "binomial",
  fish_covars = fish_vars
)

pred_pres_b <- predict(
  final_BPUE_pres,
  type    = "response",
  exclude = exclude_if_RE(final_BPUE_pres)
)
obs_pres_b  <- BData$BPUE_pos
roc_obj_b   <- pROC::roc(obs_pres_b, pred_pres_b)
auc_val_b   <- as.numeric(pROC::auc(roc_obj_b))
roc_df_b    <- data.frame(
  tpr = roc_obj_b$sensitivities,
  fpr = 1 - roc_obj_b$specificities
)
p_BPUE_pres_roc <- ggplot(roc_df_b, aes(fpr, tpr)) +
  geom_path() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  annotate(
    "text",
    x = 0.6,
    y = 0.1,
    label = paste0("AUC = ", round(auc_val_b, 3))
  ) +
  labs(x = "False positive rate", y = "True positive rate")

BPUE_panel_presence <- patchwork::wrap_plots(
  list(
    as_plot(p_BPUE_pres_smooths),
    as_plot(p_BPUE_pres_param),
    as_plot(p_BPUE_pres_roc)
  ),
  ncol = 3)

BPUE_panel_presence

# ---------- Positives component (Gamma, log link)

Bpos <- dplyr::filter(BData, BPUE_pos == 1L)

pred_fixed_BPUE_pos <- setdiff(names(Bpos), c("Weighted_BPUE_Kōura", id))

vif_BPUE_pos <- remove_high_vif_glmmTMB(
  Bpos,
  "Weighted_BPUE_Kōura",
  pred_fixed_BPUE_pos,
  threshold    = vif_thresh,
  protect_vars = character(0)
)
kept_fixed_BPUE_pos <- vif_BPUE_pos$predictors

vif_BPUE_pos$report$start
vif_BPUE_pos$report$removed
vif_BPUE_pos$report$final_vif

vars_step_BPUE_pos <- c(kept_fixed_BPUE_pos, if (INCLUDE_RE) id)

rhs_BPUE_pos <- paste(
  vapply(
    vars_step_BPUE_pos,
    function(v) build_smooth_gam(v, Bpos, custom_k, include_re_for_LID = INCLUDE_RE),
    character(1)
  ),
  collapse = " + "
)

full_ml_BPUE_pos <- mgcv::gam(
  as.formula(paste("Weighted_BPUE_Kōura ~", rhs_BPUE_pos)),
  data   = Bpos,
  family = Gamma(link = "log"),
  method = "ML",
  select = TRUE
)

remaining_BPUE_pos <- vars_step_BPUE_pos
protected_step_vars_bpos <- if (INCLUDE_RE) id else character(0)

repeat {
  rhs_now <- paste(
    vapply(
      remaining_BPUE_pos,
      function(v) build_smooth_gam(v, Bpos, custom_k, include_re_for_LID = INCLUDE_RE),
      character(1)
    ),
    collapse = " + "
  )
  m_now <- mgcv::gam(
    as.formula(paste("Weighted_BPUE_Kōura ~", rhs_now)),
    data   = Bpos,
    family = Gamma(link = "log"),
    method = "ML",
    select = TRUE
  )
  sm <- summary(m_now)

  ps <- c()
  if (!is.null(sm$p.table) && nrow(sm$p.table) > 0) {
    pcol <- intersect(colnames(sm$p.table), c("Pr(>|t|)", "Pr(>|z|)"))[1]
    pvec <- sm$p.table[, pcol]
    ps   <- c(ps, pvec[names(pvec) != "(Intercept)"])
  }
  if (!is.null(sm$s.table) && nrow(sm$s.table) > 0) {
    pvec <- sm$s.table[, "p-value"]
    names(pvec) <- rownames(sm$s.table)
    ps   <- c(ps, pvec)
  }
  drop_candidates <- ps[ps > p_cutoff_ml]
  if (!length(drop_candidates)) { red_ml_BPUE_pos <- m_now; break }

  ordered <- names(sort(drop_candidates, decreasing = TRUE))
  ordered_vars <- vapply(
    ordered,
    function(x) if (grepl("^s\\(", x))
      sub("^s\\(([^,]+).*\\)$", "\\1", x) else x,
    character(1)
  )

  # protect LID in positive BPUE model
  ordered_vars <- setdiff(ordered_vars, protected_step_vars_bpos)
  if (!length(ordered_vars)) { red_ml_BPUE_pos <- m_now; break }

  remove_v <- ordered_vars[1]
  remaining_BPUE_pos <- setdiff(remaining_BPUE_pos, remove_v)
  if (!length(remaining_BPUE_pos)) { red_ml_BPUE_pos <- m_now; break }
}

# ML comparison: FULL vs REDUCED
summary(full_ml_BPUE_pos)
summary(red_ml_BPUE_pos)
print(AIC(full_ml_BPUE_pos, red_ml_BPUE_pos))
suppressWarnings(print(anova(full_ml_BPUE_pos, red_ml_BPUE_pos, test = "Chisq")))

# Clean test of LID random effect for positive BPUE
LID_test_BPUE <- if (INCLUDE_RE) {
  m_withLID_bpos <- red_ml_BPUE_pos
  tl_bpos <- attr(terms(red_ml_BPUE_pos), "term.labels")
  tl_bpos_noLID <- tl_bpos[!grepl("LID", tl_bpos)]
  rhs_bpos_noLID <- if (length(tl_bpos_noLID)) paste(tl_bpos_noLID, collapse = " + ") else "1"
  f_bpos_noLID <- as.formula(paste("Weighted_BPUE_Kōura ~", rhs_bpos_noLID))

  m_noLID_bpos <- mgcv::gam(
    f_bpos_noLID,
    data   = Bpos,
    family = Gamma(link = "log"),
    method = "ML",
    select = TRUE
  )

  anova(m_noLID_bpos, m_withLID_bpos, test = "Chisq")
}

# Final REML
final_BPUE_pos <- mgcv::gam(
  formula(red_ml_BPUE_pos),
  data   = Bpos,
  family = Gamma(link = "log"),
  method = "REML",
  select = TRUE
)
summary(final_BPUE_pos)
mgcv::concurvity(final_BPUE_pos)
gam.check(final_BPUE_pos)

# Plots (positive BPUE) – updated layout, no LID

base_theme_bw <- theme_classic() +
  theme(
    text = element_text(family = "Arial", size = 8),
    axis.title = element_text(face = "plain"),
    axis.text = element_text(face = "plain"),
    plot.title = element_text(face = "plain"),
    strip.text = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3)
  )

clean_smooth_title <- function(x) {
  x <- gsub("^s\\((.*)\\)$", "\\1", x)
  x <- gsub("_", " ", x)
  x
}

# Smooth partial effects (exclude LID explicitly)
sm_bpue <- smooth_estimates(final_BPUE_pos)
bpue_smooth_names <- setdiff(unique(sm_bpue$.smooth), "s(LID)")
bpue_smooth_plots <- lapply(bpue_smooth_names, function(s) {
  gratia::draw(final_BPUE_pos, select = s, scales = "fixed", se = FALSE) +
    labs(title = clean_smooth_title(s), x = NULL, y = "Partial effect") +
    base_theme_bw
})

p_BPUE_pos_smooths <- patchwork::wrap_plots(bpue_smooth_plots, ncol = length(bpue_smooth_plots)) +
  patchwork::plot_annotation(tag_levels = "a")

# Scatter diagnostics (population-level prediction, no LID)
pred_pos_b <- predict(final_BPUE_pos, type = "response")
obs_pos_b  <- final_BPUE_pos$y
R2_b   <- cor(pred_pos_b, obs_pos_b, use = "complete.obs")^2
RMSE_b <- sqrt(mean((pred_pos_b - obs_pos_b)^2, na.rm = TRUE))

df_scat_b <- data.frame(Predicted = pred_pos_b, Observed = obs_pos_b)

p_BPUE_pos_scatter <- ggplot(df_scat_b, aes(Predicted, Observed)) +
  geom_point(size = 1.8, colour = "black") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black", linewidth = 0.4) +
  annotate("text",x = max(df_scat_b$Predicted, na.rm = TRUE) * 0.7,
    y = max(df_scat_b$Observed,  na.rm = TRUE) * 0.8,
    label = paste0("R\u00B2 = ", round(R2_b, 2), "\nRMSE = ", round(RMSE_b, 2)),
    size = 3) +
  labs(x = "Predicted BPUE", y = "Observed BPUE", tag = "b") +
  base_theme_bw

# Layout: flatter top, taller bottom
BPUE_panel_positives <- (p_BPUE_pos_smooths / p_BPUE_pos_scatter) +
  patchwork::plot_layout(heights = c(1, 2))

BPUE_panel_positives

ggsave(filename = file.path(fig_dir, "BPUE_panel_positives.png"), plot = BPUE_panel_positives, width = 100, height = 120, units = "mm", dpi = 1200)
```

# Models output
```{r fig.width=5, fig.height=4, dpi=100, echo=FALSE, message=FALSE, warning=FALSE}
# OCC ML
summary(full_ml_occ)
summary(red_ml_occ)
AIC(full_ml_occ, red_ml_occ)
suppressWarnings(print(anova(full_ml_occ, red_ml_occ, test = "Chisq")))
performance::check_collinearity(red_ml_occ)
LID_test_occ


# OCC Final REML
final_occ <- mgcv::gam(formula(red_ml_occ), data = PData, family = binomial(link="logit"), method = "REML", select = TRUE)
summary(final_occ)
mgcv::concurvity(final_occ)
gam.check(final_occ)



# CPUE pos
summary(full_ml_cpue_pos)
summary(red_ml_cpue_pos)
print(AIC(full_ml_cpue_pos, red_ml_cpue_pos))
suppressWarnings(print(anova(full_ml_cpue_pos, red_ml_cpue_pos, test = "Chisq")))
performance::check_collinearity(red_ml_cpue_pos)
LID_test_cpue 

# CPUE Final REML
final_cpue_pos <- mgcv::gam(formula(red_ml_cpue_pos), data = Cpos, family = Gamma(link="log"), method = "REML", select = TRUE)
summary(final_cpue_pos)
mgcv::concurvity(final_cpue_pos)
gam.check(final_cpue_pos)




# BPUE pos
summary(full_ml_BPUE_pos)
summary(red_ml_BPUE_pos)
print(AIC(full_ml_BPUE_pos, red_ml_BPUE_pos))
suppressWarnings(print(anova(full_ml_BPUE_pos, red_ml_BPUE_pos, test = "Chisq")))
performance::check_collinearity(red_ml_BPUE_pos)
LID_test_BPUE

# BPUE Final REML
final_BPUE_pos <- mgcv::gam(formula(red_ml_BPUE_pos), data = Bpos, family = Gamma(link="log"), method = "REML", select = TRUE)
summary(final_BPUE_pos)
mgcv::concurvity(final_BPUE_pos)
gam.check(final_BPUE_pos)



# GAM results tables
gam_results_table <- function(model, model_name = deparse(substitute(model)), digits = 3) {
  sm <- summary(model)

  # ---------- Parametric table ----------
  ptab <- as.data.frame(sm$p.table)
  ptab <- rownames_to_column(ptab, "term")
  ptab <- as_tibble(ptab)

  # Identify p-value column robustly
  p_col <- names(ptab)[grepl("^Pr\\(>\\|", names(ptab))]
  if (length(p_col) != 1) {
    stop("Could not uniquely identify p-value column in p.table. Found: ", paste(p_col, collapse = ", "))
  }

  # Identify statistic column robustly (t value / z value)
  stat_col <- names(ptab)[grepl("value$", names(ptab))]
  if (length(stat_col) != 1) {
    stop("Could not uniquely identify statistic column in p.table. Found: ", paste(stat_col, collapse = ", "))
  }

  param_tbl <- ptab |>
    rename(
      estimate  = Estimate,
      std_error = `Std. Error`,
      statistic = all_of(stat_col),
      p_value   = all_of(p_col)
    ) |>
    mutate(
      component = "parametric",
      model = model_name
    ) |>
    select(model, component, term, estimate, std_error, statistic, p_value) |>
    mutate(across(where(is.numeric), ~ round(.x, digits)))

  # ---------- Smooth table ----------
  stab <- as.data.frame(sm$s.table)
  stab <- rownames_to_column(stab, "term")
  stab <- as_tibble(stab)

  # mgcv usually uses: edf, Ref.df, F or Chi.sq, p-value
  # Make robust picks
  refdf_col <- names(stab)[grepl("^Ref\\.df$", names(stab))]
  edf_col   <- names(stab)[grepl("^edf$", names(stab))]
  pval_col  <- names(stab)[grepl("^p\\-value$", names(stab))]
  stat_col2 <- names(stab)[grepl("^(F|Chi\\.sq)$", names(stab))]

  if (length(edf_col) != 1) stop("Could not uniquely identify edf column in s.table.")
  if (length(refdf_col) != 1) stop("Could not uniquely identify Ref.df column in s.table.")
  if (length(pval_col) != 1) stop("Could not uniquely identify p-value column in s.table.")
  if (length(stat_col2) != 1) stop("Could not uniquely identify test statistic column (F/Chi.sq) in s.table.")

  smooth_tbl <- stab |>
    rename(
      edf = all_of(edf_col),
      ref_df = all_of(refdf_col),
      statistic = all_of(stat_col2),
      p_value = all_of(pval_col)
    ) |>
    mutate(
      component = "smooth",
      model = model_name
    ) |>
    select(model, component, term, edf, ref_df, statistic, p_value) |>
    mutate(across(where(is.numeric), ~ round(.x, digits)))

  # ---------- Combined (paper-friendly) ----------
  combined_tbl <- bind_rows(
    param_tbl |>
      mutate(edf = NA_real_, ref_df = NA_real_) |>
      select(model, component, term, estimate, std_error, edf, ref_df, statistic, p_value),
    smooth_tbl |>
      mutate(estimate = NA_real_, std_error = NA_real_) |>
      select(model, component, term, estimate, std_error, edf, ref_df, statistic, p_value)
  )

  list(
    parametric = param_tbl,
    smooth = smooth_tbl,
    combined = combined_tbl
  )
}

# Create tables for your three final models
occ_tabs  <- gam_results_table(final_occ,      "Occupancy", digits = 3)
cpue_tabs <- gam_results_table(final_cpue_pos, "CPUE (pos)", digits = 3)
bpue_tabs <- gam_results_table(final_BPUE_pos, "BPUE (pos)", digits = 3)

occ_table  <- occ_tabs$combined
cpue_table <- cpue_tabs$combined
bpue_table <- bpue_tabs$combined

# Optional significance stars (remove if journal dislikes)
add_sig <- function(df) {
  df |>
    mutate(
      sig = case_when(
        is.na(p_value) ~ "",
        p_value < 0.001 ~ "***",
        p_value < 0.01  ~ "**",
        p_value < 0.05  ~ "*",
        p_value < 0.1   ~ ".",
        TRUE ~ ""
      )
    )
}

occ_table  <- add_sig(occ_table)
cpue_table <- add_sig(cpue_table)
bpue_table <- add_sig(bpue_table)


# Export
write.csv(occ_table,  file = file.path(fig_dir, "table_gam_occupancy.csv"), row.names = FALSE)
write.csv(cpue_table, file = file.path(fig_dir, "table_gam_cpue_pos.csv"),  row.names = FALSE)
write.csv(bpue_table, file = file.path(fig_dir, "table_gam_bpue_pos.csv"),  row.names = FALSE)

# If you want one combined table for Supplement:
all_models_table <- bind_rows(occ_table, cpue_table, bpue_table)
write.csv(all_models_table, file=file.path(fig_dir,"table_gam_all_models.csv"), row.names = FALSE)



```


# PCA test
```{r}
# Comment Frank about the modeling:
  # If you centered (on the mean) and standardized (scaled to unit variance) the predictors you could use the parameter estimates as an effect size to say something about the importance of the predictors (i.e. weak vs strong effects).


# select shoreline habitat variables for PCA
habitat_vars <- M_C_data %>%
  dplyr::select(
    Riparian_vegetation,
    Substrate_index,
    Wood_cover,
    Overhanging_trees,
    Emergent_Native,
    Submerged_Native
  ) %>%
  dplyr::filter(complete.cases(.))

# run PCA
habitat_pca <- prcomp(habitat_vars, center = TRUE, scale. = TRUE)

# variance explained
summary(habitat_pca)

# loadings
habitat_pca$rotation

# extract PC1 scores
pca_scores <- habitat_pca$x[, 1]

# add PC1 back to main dataset
M_C_data_pca <- M_C_data %>%
  dplyr::mutate(
    Habitat_complexity_PC1 = NA_real_
  )

M_C_data_pca$Habitat_complexity_PC1[complete.cases(habitat_vars)] <- pca_scores

# Extract scores and loadings
scores <- as.data.frame(habitat_pca$x[, 1:2])
scores$Site <- rownames(scores)

loadings <- as.data.frame(habitat_pca$rotation[, 1:2])
loadings$Variable <- rownames(loadings)

# Scale loadings for plotting
loading_scale <- 3

ggplot(scores, aes(PC1, PC2)) +
  geom_point(alpha = 0.6) +
  geom_segment(
    data = loadings,
    aes(x = 0, y = 0,
        xend = PC1 * loading_scale,
        yend = PC2 * loading_scale),
    arrow = arrow(length = unit(0.2, "cm"))
  ) +
  geom_text(
    data = loadings,
    aes(x = PC1 * loading_scale,
        y = PC2 * loading_scale,
        label = Variable),
    hjust = 0.5, vjust = -0.5
  ) +
  labs(
    x = paste0("PC1 (", round(summary(habitat_pca)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(habitat_pca)$importance[2,2] * 100, 1), "%)")
  )

```